---
title: "Excel training"
execute: 
  freeze: false
listing:
  contents:
    - "excel/*.qmd"
  type: table
  categories: true
  fields: [title, categories]
---

We run two multi-session Excel courses. For Excel beginners, we run an introduction to Excel for people working in health and social care. That consists of seven linked sessions which are delivered live on Teams:

+ [Excel first steps](excel/excel_first_steps.qmd)
+ [Tidy data in Excel](excel/tidy_data_in_excel.qmd)
+ [References and names in Excel](excel/references_and_names_in_excel.qmd)
+ [Excel formatting](excel/excel_formatting.qmd)
+ [Excel tables](excel/excel_tables.qmd)
+ [Excel formulas](excel/excel_formulas.qmd)		
+ A final practical review, revision, and Q&A session

For more experienced users with some prior Excel experience, we run a skill builder course to help them develop and consoliate their skills:

+ [Excel foundations 1](excel/excel_foundations_01.qmd)
+ [Excel foundations 2](excel/excel_foundations_02.qmd) 
+ [Lookups in Excel](excel/lookups_in_excel.qmd)
+ [Excel programming](excel/excel_programming.qmd)
+ [Pivot tables and pivot charts](excel/pivot_tables_and_pivot_charts.qmd)

```{r}
source(here::here("R/bullet_list.R"))
source(here::here("R/training_sessions.R"))

cat("## Excel training schedule")
```

```{r}
KINDR::training_sessions(tr_type = "Excel", output_type = "DT")

```

## Excel session materials


