---
title: "Excel training"
execute: 
  freeze: false
listing:
  contents:
    - "excel/*.qmd"
  type: table
  categories: true
  fields: [title, categories]
editor_options: 
  chunk_output_type: console
---

We run two multi-session Excel courses. For Excel beginners, we run an introduction to Excel for people working in health and social care. That consists of seven linked sessions which are delivered live on Teams:

+ [Excel first steps](excel/excel_first_steps.qmd)
+ [Tidy data in Excel](excel/tidy_data_in_excel.qmd)
+ [References and names in Excel](excel/references_and_names_in_excel.qmd)
+ [Excel formatting](excel/excel_formatting.qmd)
+ [Excel tables](excel/excel_tables.qmd)
+ [Excel formulas](excel/excel_formulas.qmd)		
+ A final practical review, revision, and Q&A session

For more experienced users with some prior Excel experience, we run a skill builder course to help them develop and consolidate their skills:

+ [Excel foundations 1](excel/excel_foundations_01.qmd)
+ [Excel foundations 2](excel/excel_foundations_02.qmd) 
+ [Lookups in Excel](excel/lookups_in_excel.qmd)
+ [Excel programming](excel/excel_programming.qmd)
+ [Pivot tables and pivot charts](excel/pivot_tables_and_pivot_charts.qmd)

## Which should I do?

You should certainly do Excel beginners if Excel is complete new to you, or you've never successfully made an Excel spreadsheet before. Otherwise, it's largely a matter of confidence and experience. We know that most people have never had a proper introduction to Excel before, and think it's worth spending a bit of time to ensure that you're happy with how the basics work, so we tend to encourage people to complete the beginners course if they're in any doubt about their suitability.

To complete the intermediate-level Excel skill-builder course, we'd expect colleagues to have substantial practical experience of working with Excel. Specifically, we'd expect attendees to be able to confidently:

* open, save, close Excel files, and add worksheets to existing Excel workbooks
* understand how A1 referencing works
* read and write simple Excel formulas
* understand and change basic number formatting (for example, getting dates to behave)

If you can't do any of these, or you'd have to think hard about how to do them, we'd strongly suggest that you attend the beginner's course first before moving on to the intermediate-level sessions.


```{r}
# source(here::here("R/bullet_list.R"))
# source(here::here("R/training_sessions.R"))

cat("## Excel training schedule")
```

```{r}
KINDR::training_sessions(tr_type = "Excel", output_type = "DT", start_date = as.character(lubridate::today() - 1))

```

## Excel session materials


