---
title: "Schedule"
execute: 
  freeze: false
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false
knitr::opts_chunk$set(echo = F, warnings = F, message = F)
```

## Current schedule

Most KIND training happens in six-week core training blocks that repeat four times per year. Most of our training - especially the Excel and Power BI courses - will be contained in those blocks. The blocks broadly align with school terms, because that's when (empirically) we get the best attendance at training.

This year, those core blocks will be:

- in the late summer/early autumn, between 1st Sept and 10th Oct 2025
- in the winter, between 3rd Nov and 12th Dec

### Theme weeks

Theme weeks occupy a whole week with a single topic, with 10-20 events scheduled on a single topic. This year, we'll have the following theme weeks:

+ week commencing 19th May 2025 - Power Query
+ w/c 30th June 2025 - Debugging/Troubleshooting
+ w/c 25th August 2025 - Open Source / copyleft
+ w/c 27th October 2025 - No Stupid Questions


## Available training sessions:

```{r}
source(here::here("R/training_sessions.R"))

training_sessions(dt_output = TRUE) 
```

## Training menu

These are all the stand-alone training sessions currently on offer. Please get in touch if you'd like to see something additional:

```{r}
readr::read_csv(here::here("data/KIND_training_sessions.csv")) |>
  DT::datatable(filter = "top")
```
