---
editor_options: 
  chunk_output_type: console
---

## Introduction

+ plotly is the major alternative to ggplot
+ the code is longer, but easier to understand
+ the basics are better (accessibility and interactivity, especially)
+ but there's a much smaller plotly user base
+ big jump from the ggplot way of thinking
+ the main reason you're not using it already: the documentation is atrocious

## Setup

```{r}
library(plotly)
library(ggplot2) # for a bit of compare-and-contrast
library(dplyr) # bits of data wrangling
library(palmerpenguins) # sample data
```

## Starting a plot object

:::{.panel-tabset}

### plotly
```{r}
penguins |>
  plot_ly()
```

### ggplot

```{r}
penguins |>
  ggplot()
```

:::

## Adding points

Two big differences:
* plotly will infer a plausible structure if you don't specify one, while ggplot doesn't
* use `~` to show which variables map onto which axes, rather than `aes()`

:::{.panel-tabset}

### plotly
```{r}
penguins |>
  plot_ly(x = ~bill_length_mm, y = ~ bill_depth_mm) # will infer best structure
```

### ggplot

```{r}
penguins |>
  ggplot() +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm))
```

:::

## Non-spatial representations

:::{.panel-tabset}

### plotly
```{r}
penguins |>
  plot_ly(x = ~bill_length_mm, 
          y = ~ bill_depth_mm, 
          color = ~species, 
          type = "scatter") # specify desired structure
```

### ggplot

```{r}
penguins |>
  ggplot() +
  geom_point(aes(x = bill_length_mm, 
                 y = bill_depth_mm,
                 colour = species))
```

:::

## Styling

:::{.panel-tabset}

### plotly
```{r}
peng_plotly <- penguins |>
  plot_ly(x = ~bill_length_mm, 
          y = ~ bill_depth_mm, 
          color = ~species, 
          type = "scatter") |> # specify desired structure
  layout(title = "100% pure plotly penguins",
         legend = list(title = list(text = "Three penguin species")))

peng_plotly 
```

### ggplot

```{r}
peng_ggplot <- penguins |>
  ggplot(aes(x = bill_length_mm, 
                 y = bill_depth_mm,
                 colour = species)) + # move the aesthetics in case we decide to get fancy
  geom_point() + 
  ggtitle("Proudly pluralistic about ggplotting penguins") +
  labs(colour = "A penguin-type trio")

peng_ggplot
```

:::

## Extra series

:::{.panel-tabset}

### plotly
```{r}
penguins |>
  plot_ly(x = ~bill_length_mm, 
          y = ~ bill_depth_mm, 
          color = ~ species,
          type = "scatter") |> # specify desired structure
  layout(title = "100% pure plotly penguins",
         legend = list(title = list(text = "Three penguin species"))) |>
  add_histogram2dcontour(x = ~ bill_length_mm, 
          y = ~ bill_depth_mm)

  # plot_ly(x = ~ bill_length_mm, 
  #         y = ~ bill_depth_mm, 
  #         type = "histogram2dcontour") 


```

### ggplot

```{r}
peng_ggplot + 
  stat_density_2d(aes(group = species))
```

:::
