---
output: html_document
editor_options: 
  chunk_output_type: console
date: 2024-07-07
---
## Welcome!

-   this session is an 🌶🌶 **intermediate practical** designed for those with some R experience

## Session outline

+ why functions?
+ basic syntax
+ understanding how functions work
+ vectorised functions
+ the mysteries of the paired curly brackets `{{}}` and the dots `...`
+ anonymous functions
+ a bunch of ways to call functions

## Why functions?

+ most beginners write repetitious code
+ repetitious code is hard to maintain
+ functions give you an easy way of repeating chunks of code

## Basic syntax

+ think of this as a way of repeating yourself
+ in time-honoured fashion...

```{r}
hi_wrld <- function(){
  "hello world"
}

hi_wrld()
```

## Adding arguments

+ most of the time, you'll want to add **arguments** to your function
    + add a variable name inside the round bracket of `function`
    + use that variable name in your function body

```{r}
hi_wrld_n <- function(n){
  paste(rep("hello world", n))
}

hi_wrld_n(4)
```

## Another argument

+ you can add another argument
+ either position or name can be used in the function call

```{r}
hi_name_n <- function(name, n){
  rep(paste("hello", name) , n)
}

hi_name_n("sue", 4)
hi_name_n(n = 3, name = "tango") # evil but legal
```

## Defaults

```{r}
hi_name_n_def <- function(n, name = "janelle"){
  rep(paste("hello", name) , n)
}

hi_name_n_def(n = 4)
hi_name_n_def(n = 2, name = "bruce")
```

## Namespacing

If you plan to use non-base R functions in your function, it's always a good idea to namespace them:

```{r}
hp_av <- function(cyls = 4){
  mtcars |>
    dplyr::filter(cyl == cyls) |>
    dplyr::summarise(mean_hp = round(mean(hp)))
}

hp_av(8)
```

## Thinking about functions

+ R functions are ordinary R objects, usually bound to a function name
+ there are several ways of investigating functions using ordinary R code. Two ways of investigating aspects of functions we've already discussed:

```{r}
hp_av |>
  formals() # the arguments

hp_av |>
  body() # the body
```

## Operators and functions

+ most R operators are really functions too - properly referred to as infix functions. You can use them in the standard prefix way using backticks:

```{r}
`+`(2, 2)
```
If you're feeling ...experimental, you can also convert a standard function to an infix function by wrapping the function name in `%%`:

```{r}
`%tri%` <- function(a, b){
  sqrt(a^2 + b^2)
}

3 %tri% 4
```


## `class()`/`typeof()`

For user-defined functions, and functions from packages, this means `class()` gives sensible results:

```{r}
class(hp_av)
typeof(hp_av) # run away!
```
## Primitive functions

Note that's not true for some built-in functions written in C:


```{r}
typeof(sum)
```


## Environments
```{r}
hp_av |>
  environment() # the environment
```

+ Think of the environment as the space in which the function is executed
+ Much more to say really - but [*Advanced R*](https://adv-r.hadley.nz/environments.html) is the right place for that

## Scoping

+ the reason environments are important: names defined within scope mask those outwith:

```{r}
a <- 10

scope_test <- function(n) {
  a <- 5
  n + a
}

scope_test(5)
```
## Name masking

+ if names aren't defined within the environment of the function, it'll look one level up

```{r}
a <- 10

scope_test2 <- function(n) {
  # a <- 5
  n + a
}

scope_test2(5)
```
## Lazy evaluation

+ names get used at execution only. This explains the following odd behaviour:

```{r}

a <- 10

scope_test3 <- function(n) {
  # a <- 5
  a
}

scope_test3()

```
You'd usually expect the absence of `n` in the function call to produce an error. But because the body never uses `n`, no problem!

## Vectorised functions

+ most functions in R are [vectorised](https://bookdown.org/rdpeng/rprogdatascience/vectorized-operations.html)

```{r}
round(c(1.2, 3.2, 5.4, 2.7), 0)
```

+ that means that mostly, our functions will end up vectorised without us doing any work at all

```{r}
div_seven_n_round <- function(nums){
  round(nums / 7, 0)
}

numbers <- rnorm(4, 5, 50)

numbers

div_seven_n_round(numbers)
```

+ **but** there are a few cases where that can fail: most famously, using `if`/`else`

```{r}
is_even <- function(n){
  
  if(n %% 2){
  paste(n, "is odd")
} else {
  paste(n, "is even")
}
  
}
is_even(9)
is_even(10)
try(is_even(9:10))

```

## Three solutions

### vectorize with `Vectorize`

```{r}
is_even_v <- Vectorize(is_even)
is_even_v(9:10)
```

### apply

+ apply with `lapply` / `purrr::map` with `Vectorize`
```{r}
lapply(9:10, is_even)
purrr::map(9:10, is_even)
```

### refactor

+ refactor to avoid scalar functions

```{r}
is_even_rf <- function(n){
  ifelse(n %% 2, paste(n, "is odd"), paste(n, "is even"))
}
is_even_rf(9:10)
```

## `{{}}`

### What's the problem?
```{r}
mtcars |>
  dplyr::summarise(average = round(mean(hp)))
```
so

```{r}
carmo <- function(column){
  mtcars |>
    dplyr::summarise(average = round(mean(column)))
  }
```

+ but...
```{r}
try(carmo(hp))
```
> object 'hp' not found

+ we get used to R (and particularly tidyverse) helping us with some sugar when selecting column by their names
  + `mtcars$hp` / `mtcars |> select(hp)`
  + effectively, we're just able to specify `hp` like an object, and R figures out the scope etc for us
+ that misfires inside functions. R isn't sure where to look for an object called `hp`

### Enter `{{}}`

```{r}
carmo_woo <- function(column){
  mtcars |>
    dplyr::summarise(average = round(mean({{column}})))
}

carmo_woo(hp)
```

+ for 95% of purposes, take `{{}}` as a purely empirical fix
+ but, if you're very enthusiastic:
    + `{{}}` [defuses and injects](https://rlang.r-lib.org/reference/topic-metaprogramming.html) the column name
    + equivalent to `!!enquo(var)`
    
## ...

+ pass arbitrary arguments into/through a function with ...

```{r}
dotty <- function(n, ...){
  rep(paste(..., collapse = ""), n)
}

dotty(4, letters[1:5])

```

If you need to collect values from `...`, `rlang::list2()` is probably the right way to do that:


```{r}
list_out <- function(...){
  rlang::list2(...)
}

list_out(1, 4, 9)

```

## Anonymous functions

In some cases, you might find yourself creating a function that's only going to be used in a single location. In that case, it's possible to define a nameless (anonymous) function, which is concise-but-nasty. For example:

```{r}
(\(x) x * 2)(5)
```

Here:

+ `\(x)` defines an anonymous function that takes one argument `x`.
+ `x * 2` is the body of the function which determines how the function works
+ `(5)` is the value that we're supplying to the function

Most usually, that anonymous function won't be used with a single value, but within a pipe. Here, the supplied value is dropped, and substituted by the values passing along the pipe:

```{r}
mtcars |>
  dplyr::select(wt) |>
  (\(x) x * 2)() |> # Double the weight of all the values
  dplyr::slice_max(wt, n = 3)
```

### Anonymous functions and the magrittr pipe

The magrittr pipe (`%>%`) offers a simple method of avoiding the horrid anonymous function syntax using the `.` placeholder:
  
```{r}
library(magrittr)

mtcars %>%
  dplyr::select(wt) %>%
  { . * 2 } %>% # Double the weight of all the values
  dplyr::slice_max(wt, n = 3)
```

## Calling functions from lists

```{r}
funs <- list(
  half = function(x) x / 2,
  double = function(x) x * 2
)

funs$half(89)
```


## `do.call()`

+ For when you want to call a function over a stored set of arguments - e.g. where you have values held as a list:

```{r}

dfs <- list(
  a = tibble::tibble(x = 1, y = 2),
  b = tibble::tibble(x = 3, y = 4)
)

do.call(rbind, dfs)

```

## Resources
+ **best = home made!** Refactor something simple in your code today.
+ hard to beat the treatment of functions in [R4DS](https://r4ds.hadley.nz/functions)
+ the [Rlang page on data masking](https://rlang.r-lib.org/reference/topic-data-mask.html) is surprisingly sane for such a complicated area