---
author: "Brendan Clarke, NHS Education for Scotland, [brendan.clarke2@nhs.scot](mailto:brendan.clarke2@nhs.scot)"
date: "2024-06-28"
title: "Measures in Power BI"
execute:
  freeze: auto
categories: [Power BI, intermediate]
---


```{r}
#| echo: false
#| results: asis
#| fig-align: left
#| fig-height: 0.6
#| fig-width: 3

source(here::here("R/feed_block.R"))
feed_block("Measures in Power BI")

```


:::{.callout-note collapse=false appearance='default' icon=true}
## Session materials
+ [all materials {{< iconify ph:file-zip size=2x >}}](src/measures_in_power_bi.zip)
+ slides [{{< iconify ph:file-html size=2x >}} html](src/measures_in_power_bi.html) / [{{< iconify ph:file-pdf size=2x >}} pdf](src/measures_in_power_bi.pdf)
+ sample files [{{< iconify simple-icons:powerbi size=2x >}} .pbix](data/ae_weeks_start.pbix) / + [{{< iconify ph:file-xls size=2x >}}.xlsx](data/excel_measures.xlsx)
:::

## Welcome

+ this session is for 🌶🌶 intermediate users
+ this session can be followed practically - you'll need Power BI desktop ([plus the sample file](https://github.com/NES-DEW/KIND-training/raw/main/bi_training/data/ae_weeks_start.pbix)) to do that



```{r}
```


## Session outline

+ DAX
+ measures
  + Excel measures
  + measures vs calculated columns
+ basic measures for data summary
+ more complex measures for re-filtering

## DAX

+ **d**ata **a**nalysis e**x**pressions
+ based on Excel formula language
+ been in use for about 15 years, initially in Excel

## Let's try one in Excel

+ [sample Excel](https://github.com/NES-DEW/KIND-training/raw/main/bi_training/data/excel_measures.xlsx)
+ PivotTable plus data model needed
+ Measures live in tables
+ written in DAX: `=CONCATENATEX(VALUES(Table134[Fluid spilled on patients]), Table134[Fluid spilled on patients], ",")`

## Into Power BI

+ [sample file](https://github.com/NES-DEW/KIND-training/raw/main/bi_training/data/ae_weeks_start.pbix)
+ confusingly, we find DAX in two different places...

## Calculated columns

+ Let's calculate an average
+ `aver_att = AVERAGE(ae_activity[att])`

## Measures

+ dynamically recalculate depending on filter context
  + so need to live on a card
+ so same DAX = different result
+ `aver_att_meas = AVERAGE(ae_activity[att])`


## Basic functions for measures

+ `SUM`, `AVERAGE`, `COUNT`, `MAX`
+ measure-in-measure
+ `&` and `ROUND`

## Row-wise operations
+ `SUMX`
+ `CONCATENATEX` (from Excel example)

## More interesting measures: filter

+ `count_12 = COUNTROWS(FILTER(ae_activity, ae_activity[over12] > 20))`
+ `FILTER` allows us to look at how parts of our data change

## More interesting measures: calculate

+ `count_12_calc = CALCULATE(COUNT([att]), ae_activity[over12] > 20)`
+ `CALCULATE` changes the default filter context

## More interesting measures: keepfilters

+ `count_12_calc_kf = CALCULATE(COUNT([att]), KEEPFILTERS(ae_activity[over12] > 20))`
+ `KEEPFILTERS` pokes the default filter context through into a `CALCULATE`



+ [Definitive Guide to DAX](https://www.waterstones.com/book/definitive-guide-to-dax-the/marco-russo/alberto-ferrari/9781509306978)
+ [SQLBI CALCULATE](https://www.youtube.com/watch?v=HxZLkmpY6BA)
+ [SQLBI KEEPFILTERS](https://www.youtube.com/watch?v=Cj6rI0fRCaQ)

