---
title: A Data-Centric Introduction to Python
execute: 
  echo: true
  output: true
  freeze: false
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

A note on this mixed R/Python Quarto file: the text and Python sections were written in a Jupyter notebook, then converted into a Quarto document by:

`quarto convert a_data_centric_introduction_to_python.ipynb`

R sections, and the tabsets were added, and the document was then knitted using R Quarto in Rstudio by attaching the reticulate [@reticulate] package:

```{r}
#| echo: true
library(reticulate)
```

```{r}
#| echo: false

if (.Platform$OS.type != "windows") {
  reticulate::use_virtualenv("r-reticulate", required = TRUE)
} else {
  use_python("C:/Users/brendancl/AppData/Local/r-miniconda/envs/r-reticulate/python.exe")
}

```

# A data-centric introduction to Python

This is a friendly beginner session introducing users to Python. It's health-and-social-care opinionated, assumes no previous Python knowledge, and will have lots of scope for practical demonstrations. Given that lots of users in the KIND network will have some prior experience of R, we'll introduce some key Python features by comparison with R

# Session structure

-   a brief general-purpose chat for intro to the language
-   how to read and write Python (jupyter/VS Code/Posit workbench/positron)
-   a side-note about Excel Python
-   Python for R developers - a practical demonstration

## Python introduction

-   ["Python is a high-level, general-purpose programming language."](https://en.wikipedia.org/wiki/Python_(programming_language))
-   massive user-base
-   highly extensible and flexible ($10^5$ modules)\
-   the [second-best language for everything](https://www.youtube.com/watch?v=OkuGfa3zcL8)
-   multi-paradigm (oop, structured, ...)

# Reading and writing Python

-   you'll need:
    -   Python, [currently at 3.12](https://docs.python.org/release/3.12.4/)
    -   (almost certainly) something to manage modules - like [pip](https://pypi.org/project/pip/) or [conda](https://conda.io/projects/conda/en/latest/user-guide/install/index.html)
    -   (almost certainly) an integrated development environment. Loads of options:
        -   practical demo of [Jupyter labs](https://jupyter.org/)
        -   non-free use in [posit.cloud](https://posit.cloud)
        -   Rstudio via [reticulate](https://cran.r-project.org/web/packages/reticulate/index.html) / Jupyter
        -   [VSCode](https://code.visualstudio.com/), which is pretty well industry standard for the wider Python ecosystem
        -   [positron](https://github.com/posit-dev/positron), which is the new kid for data-flavoured Python work

## Excel Python

-   Python is coming to Excel, apparently...
    -   roll-out slower than expected
-   gives an alternative to VBA etc
-   code gets executed in the cloud, so no infrastructure faff...
    -   but a potential information governance headache
-   on the offchance that you have it available, `=PY()` is the key function
    -   see ["hello world!"-type walkthrough](https://support.microsoft.com/en-gb/office/get-started-with-python-in-excel-a33fbcbe-065b-41d3-82cf-23d05397f53d)

## Python for R people

You're welcome to follow along using the free basic Python set-up at [W3schools](https://www.w3schools.com/python/trypython.asp?filename=demo_default)

-   "hello world!"
-   indents vs brackets
-   Rmarkdown vs Jupyter
-   packages vs modules - for data from csv comparison
-   basic work with tabular data - for methods
-   vector/tibble/list vs list/tuple/dict/set - for vectorisation vs list comps
-   pandas for tabular data
-   plotting comparison

## "hello world!"

Initially, there's very little to choose between R and Python, and everything is likely to feel very familiar..

::: panel-tabset
### Python

```{python}
print("hello world!")

1 + 2

hw = "hello " + "world" + "!"
hw
```

### R

```{r}
"hello world"

1 + 2

hw <- paste("hello", "world", "!")
hw

```
:::

## Indents

-   a first big difference: indents *matter* in Python
-   they're non-optional with proper syntactic function
    -   broadly correspond to curly brackets in R

::: panel-tabset
### Python

```{python}
word = "care"

if word == "care":
    print("I have found someone from care")
else:
    print("No, I haven't found anyone from care")
```

### R

```{r}
word <- "care"

if (word == "care") {
  print("I have found someone from care")
} else {
  print("No, I haven't found anyone from care")
}

```
:::

## Rmarkdown/Quarto vs Jupyter

-   Jupyter provides interactive code- and markdown editing. Compare to the render/knit-based workflow of qmd/Rmd
-   web-based, so perhaps more like posit.cloud / workbench than Rstudio
-   comparatively harder to edit .ipynb files than .Rmd/.qmd in other tools

## Packages vs modules

We'll load the [pandas](https://pandas.pydata.org/docs/reference/api/pandas.Index.html) module in Python, and the readr package in R [@readr] to compare and contrast loading external functions. We'll use those to read some sample data (the KIND book of the week dataset).

::: panel-tabset
### Python

```{python}
botw_dat = "https://raw.githubusercontent.com/NES-DEW/KIND-community-standards/main/data/KIND_book_of_the_week.csv"

import pandas
botw = pandas.read_csv(botw_dat)

```

But we also have a lot of options for loading modules. We can alias, most usefully to give us short names for commonly-used functions:

```{python}
import pandas as pd
botw = pd.read_csv(botw_dat)
```

We could even load an individual function from a module:

```{python}
from pandas import read_csv as read_csv
botw = read_csv(botw_dat)

# one minor bit of cheating - we'll coerce the Year column to numeric
botw = botw.replace("1979 (1935)", 1979)
botw["Year"] = pandas.to_numeric(botw["Year"])
```

### R

There are comparatively fewer options for package loading in R. You'd traditionally attach a whole package using `library`:

```{r}
library(readr)

botw_dat <- "https://raw.githubusercontent.com/NES-DEW/KIND-community-standards/main/data/KIND_book_of_the_week.csv"

botw <- read_csv(botw_dat)
```

You can load individual functions by namespacing via `::`:

```{r}
botw <- readr::read_csv(botw_dat)
```

It is also possible, although non-standard, to alias individual functions:

```{r}
steve <- readr::read_csv
botw <- steve(botw_dat)
```
:::

## Fun with tabular data

Doing some basic playing with our tabular data shows that Python uses methods - like a local version of a functions that are specific to certain types of object. While methods [can be used in R](https://stat.ethz.ch/R-manual/R-devel/library/methods/html/Methods_Details.html), in practice most R code relies on functions.


::: panel-tabset
### Python

Both [`shape`](https://pandas.pydata.org/docs/reference/api/pandas.DataFrame.shape.html) and [`index`](https://pandas.pydata.org/docs/reference/api/pandas.Index.html) are methods that we've imported from pandas. They'll only work in pandas objects, which we'll talk about more below.

```{python}
botw.shape # shape is a method
len(botw.index) # as is index
botw.shape[0] # Python is 0-indexed
```

### R

```{r}
dim(botw) # dim is a function
nrow(botw) # as is nrow
dim(botw)[1] # Python is 1-indexed
```
:::

## Data types

::: panel-tabset
### Python

-   there are four basic data types in Python
  - list
  - tuple
  - dict
  - set

```{python}
numbers_list = [1,2,3,4,5] # changeable
numbers_list
numbers_tuple = (1,2,3,4,5) # unchangeable
numbers_tuple
numbers_dict = {"one":1, "two":2, "three":3} # changeable (now), no duplicates
numbers_dict
numbers_set = {1,2,3,4,5} # unchangeable, no duplicates
numbers_set
```

```{python}
# Modify in place semantics
numbers_list.reverse()
```

### R

-   R has [several basic data types](http://adv-r.had.co.nz/Data-structures.html), but in practice only three are commonly encountered. These are the vector, the data frame, and the list (confusing!):

```{r}
numbers_vector <- c(1,2,3,4,5)
numbers_vector
numbers_dataframe <- data.frame(nums = numbers_vector)
numbers_dataframe
numbers_list <- list(numbers_vector, numbers_dataframe)
numbers_list
```
:::

## Loops, list comprehensions, and vectorization

There are various methods for repeatedly running code. We'll demonstrate a couple of simple methods here. Note that both Python and R have rich and powerful functional programming tools available (like `map`), but we'll park those for now.

::: panel-tabset
### Python

You'll need to use loops, or (much nicer) list comprehension in Python. There's no exact counterpart of R's vectorized functions:

```{python}
double_numbers_loop = []

for n in numbers_list:
    double_numbers_loop.append(n * 2)

double_numbers_loop
```

## List comprehension

Like a lovely lightweight loop syntax

```{python}
double_numbers_list =  [n*2 for n in numbers_list]
double_numbers_list

# and, more fancy...
double_even_numbers_list =  [n*2 for n in numbers_list if (n%2 == 0) ]
double_even_numbers_list
```

### R

By and large, R is at its best with vectorized functions:

```{r}
double_numbers_vector <- numbers_vector * 2
double_numbers_vector
```

Loops are possible too

```{r}
double_numbers_loop <- vector("numeric", length = length(numbers_vector))

for (i in numbers_vector) {
  double_numbers_loop[i] <- i * 2
}

```

R has [copy-on-modify semantics](https://stackoverflow.com/questions/15759117/what-exactly-is-copy-on-modify-semantics-in-r-and-where-is-the-canonical-source), and so care needs to be taken to avoid writing poorly-performing loops. That means that loops are used comparatively rarely in R.
:::

## Tabular data basics

::: panel-tabset
### Python

-   we'll do a quick overview of pandas, based on [their excellent 10 minute overview](https://pandas.pydata.org/docs/user_guide/10min.html)
-   our `botw` object is a DataFrame, which is based on a dict
    -   like tibbles, DataFrames can contain columns of different types

```{python}
botw.dtypes # find out what we're dealing with
botw.head() # shows first few rows
botw.index # effectively counts rows
botw.columns # gives column names
botw.describe() # simple summary
botw.sort_values("Year") # sorting by column values
botw["Date"] # selecting a column and creating a series
botw[2:4] # subsetting by index using a slice and returning a DataFrame
botw[["Date"]] # subsetting entire columns
botw.loc[4] # subsetting by index using a slice and returning a series
botw.loc[4, ["Author", "Year"]] # subsetting by index and columns and returning a DataFrame
botw[botw["Year"] > 2010].sort_values("Year") # subsetting by years, and sorting
botw[botw["Author"].isin(["Katrine Marçal", "Caroline Criado Perez"])] # finding matching values
botw.dropna() # removes any missing values in the whole DataFrame
botw["Title"].str.lower() # returning the title column as a lower-case series
botw["Date"] = pandas.to_datetime(botw["Date"],format='%d/%m/%Y') # fixing publication dates
botw.groupby(pd.DatetimeIndex(botw['Date']).month)[["Year"]].mean("Year") # average year of publication by month of botw
```

### R

```{r}
library(dplyr) # we'll need dplyr for this work
str(botw) # shows data types etc
head(botw) # shows first few rows
nrow(botw) # counts rows
names(botw) # column names
summary(botw)

botw |>
  arrange(Year) # native pipe operator in R. Piped code in Python requires modules

botw$Date # selecting a column as a vector

botw |>
  slice(3:4) # subsetting by index using slice and returning a tibble Note different indexing behaviour

botw |>
  select(Date) # subsetting entire columns

as.character(botw[5,]) # subsetting by index and coercing to a vector. This is pretty non-idiomatic in R

botw[5,] |>
  select(Author, Year) # subsetting by index and columns and returning a tibble

botw |>
  filter(Year > 2010) |>
  arrange(Year) # subsetting by filtering years, then sorting using dplyr

botw[which(botw$Author %in% c("Katrine Marçal", "Caroline Criado Perez")),] # finding matching values using base R

botw |>
  tidyr::drop_na() # removes any missing values in the whole tibble

botw$Title |>
  tolower() # returning the title column as a lower-case vector

botw |>
  mutate(Date = lubridate::dmy(Date)) |> # fixing publication dates
  group_by(month = lubridate::floor_date(Date, unit = "month")) |>
  summarise(mean_year = mean(as.numeric(Year), na.rm = T)) # average year of publication by month of botw

```
:::

## Plots

::: panel-tabset
### Python

Using [matplotlib](https://matplotlib.org/3.1.1/index.html)

```{python}
import matplotlib.pyplot as plt

plt.hist(botw["Year"], bins = [1970, 1980, 1990, 2000, 2010, 2020])

plt.title("The KIND network BotW is biased towards newer books")
```

### R

```{r}
library(ggplot2)

botw |>
  mutate(Year = readr::parse_number(Year)) |>
  ggplot() +
  geom_histogram(aes(x = Year), fill="#1F77B4", binwidth = 10, center = 1985) +
  ggtitle("The KIND network BotW is biased towards newer books") +
  theme_minimal()


```

:::
