---
date: today
execute: 
  echo: false
format:
  # typst: 
  #   papersize: a4
  html:
    fig-height: 3
    fig-width: 7
draft: true
output: markup
categories: [data literacy, beginner]
params:
  sl_date: 2026-03-25
  fn: An introduction to data
  id: KS30
title: An introduction to data
editor_options: 
  chunk_output_type: console
---

```{r}
library(readr)
library(dplyr)
library(ggplot2)
```

:::{.callout-warning collapse=false appearance='default' icon=true}
## Warning!
- This is an early-stage draft of a complex session
- Structure matters for this sort of course, and the structure is as yet unsettled
- Please treat this material as completely fallible and potentially misleading
- Not even sure what we should call it...
:::


# Current sessions overview

This is shaping up to be a seven-session beginners course in data work.

1. An introduction via a simple data journey
1. Distributions
1. Datafy your system
1. Measure, collect, and structure
1. Summarise and compare
1. Visualise and communicate
1. Going the other way: interpreting other people's data

{{< include dl_to-do.qmd >}}


{{< include dl_about.qmd >}}


{{< include dl_s01.qmd >}}

{{< include dl_s02.qmd >}}

{{< include dl_s03.qmd >}}

{{< include dl_s04.qmd >}}

{{< include dl_s05.qmd >}}

{{< include dl_s06.qmd >}}

{{< include dl_s07.qmd >}}






# Acknowledgements

Ian Douglas (NHS Highland) for basically summing up the whole course in an email. Maria McAllister (HIS) and Kristi Long (NES) for changing the way we defined data in this course. Kenneth Mack (HIS) for probing on the difference between data and information. Gail Gibbs (NHS Lanarkshire) for starting off the five questions about data line of work. Judi Evans (PHS) for overall feedback. Catherine McGrenera (NHS 24) for pressing the importance of storytelling with data. Zoe Turner (Strategy Unit) for showing why we need to start with data-fying. Ailsa King (NHS GGC) and John Rankin (PHS) for thinking about communication.

# Glossary

```{r}
#| echo: false
#| output: asis

glossary <- yaml::read_yaml("../glossary.yml")

glossary <- glossary[order(names(setNames(glossary, glossary)))] # order entries alphabetically

for (term in names(glossary)) {
  cat("### ", term, "\n\n", glossary[[term]], "\n\n", sep = "")
}
```

