---
format: html
execute: 
  output: markup
editor_options: 
  chunk_output_type: console
---

```{r}
here::i_am("skills/dl_s01.qmd")

library(readr)
library(dplyr)
library(stringr)
library(ggplot2)
library(here)
```


# An introduction via a simple data journey (session 1)
  
:::{.callout-important collapse=true appearance='default' icon=true}
## Outline blob

In this session, we'll work through a simple data task. We'll use a small dataset about beds in care homes. We'll talk about how to approach the task, demonstrate how that data should be tidied, show some examples of summarising data to help understand and communicate key findings, draw some graphs to represent our data, and talk about how this dataset might be useful for explanation, prediction, and control.

### Sentence aim

To introduce the core aspects of data work to engage participants, and to set up later sessions in this course.

### Exercises

+ E1: take an imperfect dataset and tidy it
+ E2: think about how you might have made this data, and how that might change how you use it
+ E3: practical about summarising the data in different ways
+ E4: thinking about different ways of visualising this data
+ E5: thinking about purposes: how might different summaries and visuals be useful to you
+ E6: homework - take an extended dataset

### Data

Care inspectorate data. Take a subset of c.20 care homes, include name, bed count, postcode, health board, urban/rural

### Key concepts

+ understanding the different steps in data tasks
+ thinking about explanation, prediction, and control
:::


## Introduction

What is data, how does it help us do good work in health and care, and what do we need to know about data to do that work?

In this session, we'll explore the idea that data is an especially powerful way of understanding how systems in health and care, because data is an excellent way of answering some of the most important questions about our work, such as:

+ "is our hospital busier than it was last year?"
+ "do we have proportionally more children than average in our town?"
+ "how many people are currently homeless in Scotland"

This course will give an introduction to answering these questions using data. For the purposes of this course, we'll think about data as **structured information, usually numerical, that tells us about part of a system**. Specific examples of datasets that might help us answer our three questions above:

+ a month-by-month count of patients visiting our hospital for the last two years
+ census data, showing demographics of different towns in Scotland
+ survey data estimating how many people are currently homeless nationally

One of the most important ideas in this whole course is the **data journey**. This is the idea that - if we want to answer a question about our system using data - we have to take several linked steps, in some kind of order, that let us travel from an initial hunch or query to a finished answer. Very roughly, the steps in the data journey are:

+ to try and data-fy our system, and to collect data from it
+ to take that data, and tidy it into the most helpful shape for our project
+ to summarise and visualise that data
+ to communicate those findings to others

The sessions of this course are based on steps in the data journey in more detail. But the plan for today is simpler: we'll try and complete a whole data journey, a simple one, in this session. The idea here is to give you a high-level map of what we're going to talk about later in this course. Because some of these steps can be quite technical and detailed, it's really important to keep an eye on the prize - being able to put your data to work - and the best way to do that is to work through everything once quickly.

To make today's data journey possible, we'll do a bit of cheating by starting with a carefully-chosen (and polished) dataset. Don't worry too much about what we're missing by taking this shortcut, because in the next couple of sessions we'll spend a lot of time on the ins-and-outs of building lovely datasets from real systems. 

With our data in hand, we'll then think about how to understand, summarise, visualise, and compare data that we've collected. Finally, we'll talk about how to take those statistics, and use them to do things. As we'll see, statistical summaries of data are especially helpful in helping us understand the systems we're working on. We'll look specifically at three areas where understanding is especially important: **explaining** how systems work, **predicting** what will happen as our system changes, and **controlling** the behaviour of that system.

Together, we'll describe this linked set of practices and skills as **data literacy**, which describes a set of knowledge and skills across the data journey from raw data to finished project. In this course, we'll think about data literacy as being made up of a a combination of:

+ understanding how systems can be data-fied
+ having some awareness of the tools and rules that we should use when gathering data
+ knowing a bit about some statistical tools used to understand, summarise, and compare that data
+ understanding some effective strategies for using this data (and statistics) as a basis for the explanation/prediction/control of systems, including how to communicate your findings to others
+ finally, figuring out how to use that knowledge of what you want to do with your data to simplify and streamline other aspects of the data journey



## This session

In this session, we'll do a miniature data journey. To do that, we'll work through a series of practical exercises, with discussion-based interludes on the way.

## Data for this session

This session uses some publicly-available data from the [Care Inspectorate datastore](https://www.careinspectorate.com/index.php/statistics-and-analysis/data-and-analysis). Specifically, we'll use a small extract from their main [2025 end-of-year data](https://www.careinspectorate.com/images/Statistics_and_Analysis/Datastore/250331DatastoreExternal.csv) concentrating on care homes for older adults in Inverness and Nairn. We'll have a chance to look at that data more thoroughly later, but here's a short preview to help you get acquainted:

```{r}
# data wrangling code retained for interest

# full_ci_data <- read_csv("https://www.careinspectorate.com/images/Statistics_and_Analysis/Datastore/250331DatastoreExternal.csv") |>
#   filter(CareService == "Care Home Service") |>
#   filter(Subtype == "Older People") |> 
#   select(ServiceType, ServiceName, CareHome_Main_Area_of_Care, Service_Postcode, Service_town, Service_Postcode, TotalBeds) |>
#   rename_with(\(x) str_remove(x, "Service_*")) |>
#   rename(Main_area_of_care = CareHome_Main_Area_of_Care,
#          Town = town) |>
#   arrange(desc(Name))
# 
# full_ci_data |>
#   write_rds(here("skills/data/care_inspectorate_full.rds")) |>
#   writexl::write_xlsx(here("skills/data/care_inspectorate_full.xlsx"))

full_ci_data <- 
  read_rds(here("skills/data/care_inspectorate_full.rds"))

ci_towns <- function(towns = c("Inverness", "Nairn")){
  full_ci_data |>
    filter(Town %in% towns)
}

# ci_towns() |>
#   write_rds(here("skills/data/care_inspectorate.rds")) |>
#   writexl::write_xlsx(here("skills/data/care_inspectorate.xlsx"))
```

```{r}
ci <- ci_towns()

ci |>
  slice_head(n = 5) |>
  knitr::kable()
```
Hopefully that's fairly self-explanatory: we've got one row per care home, with columns for the type of home, the name, the main area for which the home provides care, the postcode and location of the home, and the total number of beds. In this session, we'll try to use this data to answer a couple of questions about the provision of care in this area. Specifically, over this session we'll see:

:::{.callout-note collapse=false appearance='default' icon=true}

## Questions for this session

+ How many care home beds are there in total in Inverness and Nairn together?
+ How do the beds available for different main areas of care compare?
+ Does Inverness or Nairn have more care home beds? What's the right way of comparing the two?
+ Is there some kind of relationship between the size of care home, and their town?

:::

The next step is to open this yourself, and have a look at the full data set.

:::{.callout-note collapse=false appearance='default' icon=true}
## Task one: inspect your data

+ download the [dataset as an Excel file](data/care_inspectorate.xlsx) to your computer
+ open it in Excel (or whichever tool you're planning to use for this session)
+ familiarise yourself with the data: you should see that we have about a dozen rows of data
+ see if there are any areas that you might want to tidy up

:::

A reminder: this course is all about using data to answer questions about health and care systems. We're using the definition of data as **structured information, usually numerical, that tells us about part of a system**. This is probably a good time to try and develop that definition a bit. Last thing first, this is data about a system. Specifically, it's care homes in Inverness and Nairn. It's narrowly focused on that topic - there's not much other information here other than the names, locations, and types of those care homes. That's typical of good data - it's not just a collection of everything that you can think of vaguely on topic, but instead should be a lean collection of the core information about one single thing. Next, that information is structured, meaning that we've organised it in a standard way. Each row (running horizontally) is about one care home. Each column (running vertically) describes one aspect of our system. So the `TotalBeds` column, for example, only contains information about the number of beds in each home, and nothing else. That column is numerical, but none of the others are in this sample data. That probably makes it a bit unusual for health and care work, but not completely exceptional.

Because this session is intended to be a quick introduction to a data journey, this dataset is lovely and tidy. That's unfortunately quite rare in the real world. Just to note a few things that we'll discuss further along our journey:

+ the data is {{< glossary "tidy data" display="tidy" >}}. That means something very specific about the way that rows and columns are used, which we'll develop more in a later session.
+ there's no missing data
+ the values in each column are consistent (so one column is all postcodes, another all numbers, etc)
+ and we seem to have standard {{< glossary coding >}} for each value
+ and the text and postcodes seem to be nice and clean, with no weird spellings, formatting, cases, etc

As we'll see, collecting data like this can be complicated. My understanding is this actual data comes from annual returns submitted to the Care Inspectorate. Just as an introduction, we'll now do a short exercise to think about how you might go about collecting data like this:

:::{.callout-note collapse=false appearance='default' icon=true}
## Task two: think about how you might have made this data, and how that might change how you use it

This is a group discussion. We'll spend a couple of minutes thinking about ways that this data might be collected. Issues to consider:

+ how would you identify individual care homes?
+ how could you be sure that you hadn't missed any care homes in the area?
+ how would you decide which establishments were in which towns?
+ how might you find out what the main area of care provided is?

:::

As that discussion (hopefully) revealed, there are lots of practical and conceptual issues underlying this sort of simple data. Those issues, and the ways we approach them, are the basis for sessions 3 and 4 of this course. They're important, even for people who aren't planning to do any data collection themselves. For example, one of the exercises that we're going to try later in this session brings in a tiny bit of new data from the [2022 census](https://www.scotlandscensus.gov.uk/), which gives us the following estimates for the population of our two towns:

| Town | Population (2022 census) |
|----|----|
| Inverness and Culloden | 65053 |
| Nairn | 9582 |

Here's one reason why understanding the route by which data was collected matters. Our care home data talks about "Inverness". But the census data talks about "Inverness and Culloden". Are they the same thing? If not, how might you need to change your population estimates to make sure we're talking about the same places in the same way as the Care Inspectorate? 

We'll leave the discussion about that issue for another session. Now though, let's introduce one of the big ideas in data literacy: summarising.

We've got data about `r nrow(ci)` different care homes in our dataset. One of the standard approaches would now be to try and boil that down to a smaller number of values - ideally one! - that tells us about these care homes. Probably the most commonly-used way of summarising would be to add all the values together, to give us a sum. Let's try that now.

:::{.callout-note collapse=false appearance='default' icon=true}
## Task three-a: sum

+ add all the values in the `TotalBeds` column
+ report your value in the chat. What does it tell you?

:::

Even though it's simple, summing values together is really important. If we're interested in comparing Inverness and Nairn together with another part of the local area, we'd have a much easier job of comparing places. The national Care Inspectorate dataset, which I'll give you to play with at the end of the session, has data for Aberdeen too, which has `r sum(ci_towns("Aberdeen")$TotalBeds, na.rm = T)` total beds - so almost exactly twice as many care home beds as Nairn and Inverness combined, which I estimated as `r sum(ci$TotalBeds, na.rm = T)`.

In miniature, this example shows probably the most important single theme in data literacy: summarising data helps us understand how things are by allowing comparisons.

Adding things up isn't the only way we can summarise. Say we wanted to try and understand how many beds a typical care home has. We could try to eyeball a typical value from our data. If we only had two or three values (say `r paste(ci$TotalBeds[1:3], collapse = ", ")`), perhaps we could do that quite effectively. I'd think that an estimate of about 40 would probably be about right for a typical value of those three figures. But even by the time we get to the dozen or so values in our dataset, that's looking more difficult:

```{r}
paste(ci$TotalBeds, collapse = ", ")
```

I'd struggle to effectively come up with a typical/average value without making a quick note somewhere and/or counting on my fingers.  Luckily, we don't need to do that, because we can take an average. There are a few different ways of taking an average, but the underlying idea is similar - tell me what a typical result in our data looks like. One way of calculating an average is to take our sum total, and divide it by the number of homes in our data. That gives us a type of average called a {{< glossary mean >}}:

$$Mean = {Sum \over n} = {621 \over 15} = 41$$

Let's check that now:

:::{.callout-note collapse=false appearance='default' icon=true}
## Task three-b: practical about summarising the data in different ways

+ take a mean of the values in the `TotalBeds` column
+ In Excel, you can use `AVERAGE()` - or you can divide your total by a `COUNTA()` of your values

:::

There are lots of other ways of summarising data, which we'll explore in later sessions, but before we move on we should talk about a refinement of summarising our data. We could, before we summarise, form the data into groups. For instance, our care homes data has a column about the main area of care that the home provides. We could sum (or mean) each of those groups separately:

```{r}
ci |>
  group_by(Main_area_of_care) |>
  summarise(Total = sum(TotalBeds),
            Mean = mean(TotalBeds)) |>
  knitr::kable()
```

That's interesting, because these sorts of grouped summaries are a mainstay when it comes to comparing different groups in our data. For our next step, let's look at a completely different way of representing our data.

```{r}
#| eval: false

# Postcodes generated using PostcodesioR - not installed in this training project, code included for interest only
purrr::map(ci$Postcode, PostcodesioR::postcode_lookup) |>
  purrr::list_rbind() |>
  dplyr::select(postcode, latitude, longitude) |>
  dplyr::distinct() |>
  readr::write_rds(here("skills/data/postcodes.rds"))

shape <- rnaturalearth::ne_countries(returnclass = "sf",scale = 10, country = "United Kingdom") |>
  sf::st_crop(xmin = -5, xmax = -3, ymin = 57, ymax =58) |>
  sf::st_transform(crs=3857)

pc <- readr::read_rds(here("skills/data/postcodes.rds")) |>
  sf::st_as_sf(coords = c("longitude","latitude"),crs=4326) |>
  sf::st_transform(crs=3857) |>
  right_join(dat, by = join_by(postcode == Postcode))

list(shape = shape, pc = pc) |>
  write_rds(here("skills/data/carehome_map.rds"))
```

```{r}
ci_post <- ci |>
  left_join(read_rds(here("skills/data/postcodes.rds")), join_by(Postcode == postcode)) 

read_rds(here("skills/data/carehome_map.rds"))$shape |>
  ggplot() +
  geom_sf() +
  theme_void() +
  geom_sf(data = read_rds(here("skills/data/carehome_map.rds"))$pc, aes(colour = Town, size = TotalBeds))
  
```


:::{.callout-note collapse=false appearance='default' icon=true}
## Task four: thinking about different ways of visualising this data



:::

```{r}
# perhaps add a percentage 
read_rds(here("skills/data/care_inspectorate.rds")) |>
  group_by(Main_area_of_care) |>
  summarise(TotalBeds = sum(TotalBeds)) |>
  mutate(percentage = scales::percent(TotalBeds / sum(TotalBeds)))

# Does Inverness or Nairn have more care home beds? What's the right way of comparing the two?
read_rds(here("skills/data/care_inspectorate.rds")) |>
  group_by(Town) |>
  summarise(TotalBeds = sum(TotalBeds))

read_rds(here("skills/data/care_inspectorate.rds")) |>
  group_by(Town) |>
  summarise(TotalBeds = sum(TotalBeds)) |>
  mutate(Population = c(65053,9582)) 

read_rds(here("skills/data/care_inspectorate.rds")) |>
  group_by(Town) |>
  summarise(TotalBeds = sum(TotalBeds)) |>
  mutate(Population = c(65053,9582)) |>
  mutate(Beds_per_10000_residents = TotalBeds + 10000 / Population)

# Is there some kind of relationship between the size of care home, and their town?
read_rds(here("skills/data/care_inspectorate.rds")) |>
  ggplot(aes(x = Town, y = TotalBeds)) +
  geom_point() +
  geom_label(aes(label = Name), nudge_x = 0.3) +
  ylim(0,NA)

read_rds(here("skills/data/care_inspectorate.rds")) |>
  group_by(Town) |>
  summarise(Average_beds = mean(TotalBeds))
```






:::{.callout-note collapse=false appearance='default' icon=true}
## Task five: thinking about purposes: how might different summaries and visuals be useful to you


:::

:::{.callout-note collapse=false appearance='default' icon=true}
## Task six: homework - take an extended dataset

Here is an [Excel file](data/care_inspectorate_full.xlsx) (or an [.rds](data/care_inspectorate_full.rds)) of the full national Care Inspectorate data in the same format as we've been using in this session. Please use it to answer some questions:



+ Is there some kind of relationship between the size of care home, and their type (private, health board, etc)?

```{r}
read_rds(here("skills/data/care_inspectorate_full.rds")) |>
  ggplot(aes(x = Type, y = TotalBeds)) +
  geom_boxplot() +
  # geom_label(aes(label = Name), nudge_x = 0.3) +
  ylim(0,NA)
```



:::





