---
title: Writing effective surveys
draft: true
output: asis
categories: [skills, beginner]
editor_options: 
  chunk_output_type: console
params:
  sl_date: 2025-11-20
  fn: Writing effective surveys
format: html
date: "`r params$sl_date`"
---

::: {.content-visible unless-format="revealjs"}

```{r}
#| echo: false
#| results: asis
#| fig-align: left
#| fig-height: 0.6
#| fig-width: 3

source(here::here("R/feed_block.R"))
feed_block(params[["fn"]])

source(here::here("R/next_sesh.R"), local = T)
next_sesh(params[["fn"]])

```
:::

::: {.content-visible when-format="revealjs"}

:::: {.content-hidden unless-meta="draft"}

## Beware!

This is a pilot version of a training session, so do please expect lumps, bumps, and inconsistencies

::::


## KIND
The KIND network is an inclusive social learning community for people working with knowledge, information, and data across health, social care, and housing in Scotland.

* [Join our Teams channel](https://forms.office.com/pages/responsepage.aspx?id=veDvEDCgykuAnLXmdF5Jmn79kl25VpJIq3eErXXCYKBUMUpENjBJOENINDRFMUlYQTlCM1RVNzRSRy4u&route=shorturl) for updates, technical support, and events (or see our [FAQ pages](https://nes-dew.github.io/KIND-community-standards/) for more info)
* [Subscribe to our mailing list](https://forms.office.com/pages/responsepage.aspx?id=veDvEDCgykuAnLXmdF5JmpopIZB9ynRJnrPUHVFccipURjM2NkZJUkhGOFlQRjQxRFhVUTgwT0UwVyQlQCN0PWcu) to receive a weekly update about our training and activities
* Visit our [training mini-site](https://nes-dew.github.io/KIND-training/) to find out about our free technical training, including Excel, Power BI, R, and SQL
* Find out more about the [digitally-enabled workforce programme](https://learn.nes.nhs.scot/53965/digitally-enabled-workforce), which operates the KIND network


## Forthcoming sessions
```{r}
KINDR::training_sessions(start_date = params[["sl_date"]], n = 5, hide_area = T)
```

:::

## Introduction

* surveys: groups of questions designed to help you understand something in a population
* quick and low-tech way of gathering information about a topic
* definitely different from 'proper' qualitative research, etc
* closed (and usually simple) questions
* not tech-specific: paper, in-person, web form...
* *usually* completed by the members of that sample directly (or their advocate) rather than by a professional

## Of note

+ local National worries about [survey responses](https://www.gov.scot/publications/understanding-survey-nonresponse-behaviours-evidence-practical-solutions/)
+ survey-logic informs big, serious, decisions
+ surveys are often extremely badly put together

## This session

+ largely tips and tricks
+ a sensible, small, toolkit of ideas to help you understand surveys
+ an opportunity to practice making elements of surveys
+ a chance to laugh at a truly atrocious survey

## The population

+ the key idea to build useful surveys: what's the population?
* population: some specified group of people - service users, general public, left-handed furriers from Dundee...
* surveys will usually be asked of a sample of a population (unlike a census, where everyone replies)

## Core survey ideas

* **topic**: what's the survey about?
* **coverage:** who can fill in the survey?
* **sample:** who actually did fill in the survey?
* **completion:** who completed the survey questions?
* **discrimination:** what did the different answers tell us about our topic?

## An example: favourite TV show

* surveys are often about complicated topics
    * if it's easy, you don't need a survey!
* we're going to start with a simple topic: **what TV programmes do people like?**

## Before we start...

Some questions for you:

* do we really understand our topic well enough to turn it into questions?
* what's the population we're interested in?
* how should we get good coverage (and what does good look like for us?)

## Exercise 1

:::{.callout-important collapse=false appearance='default' icon=true}
## Quick survey plan

+ we'll work in small groups
+ 3 minutes to sketch out a survey that:
  * investigates part of our favourite TV show **topic**
  * in a specific **population**
  * and how you're going to achieve **coverage**
+ report back in the chat

:::




## Quick wins

* be specific
* identify a specific population, and tailor the survey to them
* worry about completion (and include that in your results)
* have a plan about how to use the results (and maybe work backwards from that)

## Questions

* what questions do we need to ask people to find out about our topic?
* how do they need to be adapted to work for our population?
* what does it tell us about our topic?
* what are we trying to discriminate?
 
## Contradictory suggestions about questions

A personal rationale for these contradictory bits of advice: keeping users happy and engaged is key to writing good survey questions. It should be obvious what a question is getting at, why it's being asked.

+ suggestion: align your questions with your topic, and make it obvious. Don't leave users pondering questions like "I wonder why they're interested in that??"
+ suggestion: make your questions discriminate between different options, and make each of your questions obviously different. Avoid 29 fractionally-different questions about basically the same thing
+ suggestion: include absolutely as few questions as possible

## Why so short?

Most people have some degree of survey fatigue. Don't make that worse by writing a three-volume epic. Tread lightly on your respondent's time, and make it clear what the benefit of answering is.

+ long surveys have poor discrimination (because people get bored)
+ long surveys have poor response rates
+ let users pick options rather than generate text themselves wherever possible
+ long surveys produce sampling problems, where only the most motivated will complete. And the views of the most motivated don't reflect the views of most people.

:::{.callout-note collapse=false appearance='default' icon=true}
## Reminder
* **sample:** who actually did fill in the survey?
* **completion:** who completed the survey questions?
* **discrimination:** what did the different answers tell us about our topic?
:::


## Exercise 2

::::{.callout-important collapse=false appearance='default' icon=true}
## Design a few questions

+ again, in small groups...
+ 1 minute to review your earlier topic/population/coverage note
+ 3 minutes to sketch 2 questions to:
    + be informative about your topic
    + discriminate different views about important parts of your topic
    + but be as simple as possible
+ report back in the chat

::::


## Then, after the data collection...

* what sort of sample did we get?
  * how does that compare with our planned coverage?
  * what did our completion look like?
* analysis (non-technical!)
* what do we do with the results?

## That's a lot to think about!

+ so let's try to look at the core analytical decisions by looking at a **terrible** survey

:::{.callout-note collapse=TRUE appearance='default' icon=true}
## SURVEY QUESTIONS
+ What’s your favourite TV programme (free text)
+ Do you like comedy or drama? (exclusive choice between yes and no)
+ an un-directed Lickert-like set of responses for
    + I like longer programmes
    + I like shorter programmes
    + I have a TV
    + I am under seven feet tall
+ Do you have a second favourite TV programme? (free text)
+ How long have you watched your favourite programme for? (exclusive choice between less than a year, less than two years, more than three years)
+ Who are you? (long free text)
:::


## What did we learn?

+ even with a terrible survey, we can definitely learn something
+ for example, we've got information about people's favourite TV programmes:

```{r}
# echo: false
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)

bad <- readxl::read_xlsx(here::here("skills/data/KIND_effective_surveys_bad.xlsx"))

recodo <- list(
"What's your favourite TV programme" = "fave",                 
"Do you like comedy or drama?"  = "comedy",                          
"I like longer programmes"  = "long",                              
"I like shorter programmes"    = "short",                           
"I have a TV"  = "tv",                                           
"I am under seven feet tall"  = "seven",                            
"Do you have a second favourite TV programme?"   = "second",         
"How long have you watched your favourite programme for?" = "second_long",
"Who are you?"   = "who")

nom <- names(recodo)

long_bad <- bad |>
  select(-c(2:6)) |>
  tidyr::pivot_longer(-ID) |>
  # mutate(name = recode(name, !!!recodo, .default = "Not a scooby")) |>
  group_by(name) |>
  count(value, sort = T)

# long_bad_ind |>

long_bad |>
  filter(name == nom[1]) |>
  slice(1:7) |>
  knitr::kable(caption = "Top 7 TV programmes")

```

::: {.content-visible unless-format="revealjs"}

Note that these favourite programmes are free-text responses. In cases like these, free text (where people can reply however they choose) are really the only way that a question like this could be structured - unless you somehow have a tidy list of all the top TV programmes that people might choose. 

Free text is though much harder to analyse than yes/no choices, or numeric ratings, or similar ways of collecting information. They can also put people off if overused, and we might start to worry about sampling in a survey with lots of long free text.

This example shows a very specific gotcha: many analytical tools are case-sensitive, so "SLOW HORSES" would be grouped differently might be different from "slow horses". We'll find similar difficulties with variant punctuation too, so "Grey's Anatomy" and "Greys Anatomy" might become separated. Let's show that now in our data:

:::

::: {.content-visible when-format="revealjs"}

## What did we learn?

* free text responses
  * necessary but messy
  * can be distracting and cause dropoffs
  * careful analytics needed
  

:::
```{r}
library(stringr)

long_bad_tc <- bad |>
  select(-c(2:6)) |>
  tidyr::pivot_longer(-ID) |>
  # mutate(name = recode(name, !!!recodo, .default = "Not a scooby")) |>
  group_by(name) |>
  mutate(value = str_replace(value, "'", "")) |>
  mutate(value = stringr::str_squish(snakecase::to_title_case(value))) |>
  count(value, sort = T)
  
# long_bad_tc |>
  # filter(stringr::str_detect(value, "(?i)anatomy"))

long_bad_tc |>
  filter(name == nom[1]) |>
  slice(1:7) |>
  knitr::kable(caption = "Top 7 TV programmes")
```

Personally, I'm delighted to see Slow Horses creeping up the list. And that's a cause for serious concern: all free text will definitely need some kind of tidy-up at the analysis stage. It's hard not to favour your favourites there, and it took me a while to realise that e.g. we had "Grays Anatomy" and "Gray's Anatomy" in our data. To avoid that sort of bias, I'd be keen to try and pre-designate what text fixing you're planning to do before you collect the data (so all lower-case without punctuation, and with manual tidying-up of spelling, for example).

```{r}
n <- bad |>
  nrow() 
```

## n

::: {.content-visible unless-format="revealjs"}

n is probably the most important single number in surveys, because it tell us how many responses we received as part of our sample. So our sample consists of `r n` responses. Related, we could also estimate a response rate from that figure. Our coverage - so the number of people invited to fill in the survey - was approximately 2000 people, because that's the daily average number of active users on the KIND network at present. Dividing n by that number gives us a response rate of approximately `r scales::percent(n/2000)`. That's pretty fair for a casual survey like this which has gone out online, and many thanks to the network members - our population - for volunteering their time.

:::

::: {.content-visible when-format="revealjs"}

* n = number of responses
* allows us to estimate a response rate
  * for the horror-survey, that's `r scales::percent(n/2000)`
:::

## Discrimination

::: {.content-visible unless-format="revealjs"}

What else can we learn? Most of the questions here don't really tell us anything at all, other than our sample being really quite diligent, and keen to try and fill-in questions that made no sense at all. In other words, these answers fail to discriminate anything from anything else: we don't know what either response really means here: 🎭 which do people like best?


```{r}
long_bad_tc |>
    filter(name == nom[2])|> 
    knitr::kable()
```

Oh dear, that's not very helpful. Unfortunately, there are several similar questions where - although our long-suffering army of survey volunteers have diligently completed responses - because the responses aren't clearly specified, we're unable to understand what those users might have thought. Another example:

```{r}
long_bad_tc |>
    filter(name == nom[3]) |>
    knitr::kable()
```

This is quite an important one, because it demonstrates the (faulty) use of an important kind of preference-scale known as a Likert. That's (usually) a five-point scale that's useful for assessing agreement/disagreement. So the usual format would be to pose a question - "how do you feel about slapstick"



:::

::: {.content-visible when-format="revealjs"}

## Discrimination

+ most of the questions here don't really tell us anything at all:


```{r}
long_bad_tc |>
    filter(name == nom[2])|> 
    knitr::kable()
```

## Discrimination

+ because the responses aren't clearly specified, we're unable to understand what those respondents might have thought. Another example:

```{r}
long_bad_tc |>
    filter(name == nom[3]) |>
    knitr::kable()
```

+ note faulty use of an important kind of preference-scale known as a Likert, used for assessing agreement/disagreement

:::




## Exercise 3

::::{.callout-important collapse=false appearance='default' icon=true}
## Rework!

+ working alone this time, please come up with a better group of questions to probe these issues
+ two minutes, and then please paste them into the chat

::::

## Missing data

```{r}
long_bad_tc |>
  filter(name == nom[4]) |>
  knitr::kable()
```

Expect missing data (the `NA` values, here). You can potentially force users to answer all your questions, but that comes at the risk of potentially tempting them to send you a lot of quickly-generated gibberish.

## Drop-off

Interestingly, our respondents managed to stick with the terrible survey, and nearly everyone answered nearly every question:

```{r}
long_bad |>
  mutate(q = match(name, names(recodo))) |>
  filter(!is.na(value)) |>
  group_by(q) |>
  summarise(n = sum(n)) |>
  ggplot() +
  geom_col(aes(x = q, y = n)) +
  scale_x_continuous(breaks = 1:9) +
  xlab("Question number") +
  ylab("Number of responses")
```

That's likely to be a result of me asking people, as politely as I could, to please *please* stick with the terrible survey and answer as diligently as possible. I suspect a real survey like this one would show a much stronger drop-off towards the later questions.


## Alignment

Some of your questions should probably be used to find out about your sample. Again, think carefully about both the theme of the question, and its ability to actually distinguish what you need to know:

```{r}
long_bad_tc |>
  filter(name == nom[6]) |>
  knitr::kable()
```

::: {.content-visible when-format="revealjs"}

## Alignment

:::

In fact, from this and the other three questions using the same un-useful scale, all we see is that people irrationally prefer some points of the scale to others:

```{r}
bad |>
  select(ID, any_of(nom[3:6])) |>
  pivot_longer(-ID) |> 
  filter(!is.na(value)) |>
  ggplot() +
  geom_histogram(aes(x = value), stat = "count")
```

## Scope of responses

You should also think about the likely scope of responses, especially if you're unwise enough to ask hugely vague "who are you"-type questions

```{r}
long_bad_tc |>
  filter(name == nom[9]) |>
  slice(1:5) |>
  knitr::kable(caption = "Top five responses to the question 'Who are you'?")
```

We almost always will need to aggregate responses: this sort of question leads to impossible to aggregate responses. This question needed some honing down (e.g. into age group or occupation or similar) and replaced with a couple of closed questions. 


## Answer options

Make the alternatives as standard and coherent as possible in exactly the way that the example below isn't. I'd have thought that asking for a number of years would have been a better way of writing this question. It's worth mentioning that iteratively developing and testing questions is probably a good idea if you're doing this sort of work at larger scales.

```{r}
long_bad_tc |>
  filter(name == nom[8]) |>
  knitr::kable()
```

## Wording options

It's probably a good idea to group similar questions together. We'd also suggest trying to word similar questions as similarly as possible. For example, in our terrible survey we received quite different responses when we asked "What’s your favourite TV programme" as compared to the follow-up "Do you have a second favourite TV programme?", which received large numbers of (facetious) yes/no answers:

```{r}
long_bad_tc |>
  filter(name == nom[7]) |>
  slice(1:5) |>
  knitr::kable(caption = "Top five responses from a follow-up question about a second-favourite")
```

::: {.content-visible when-format="revealjs"}

## Wording options

:::


I'm not certain why the follow-up about second-favourite shows didn't seem to have any groups in it, but I do wonder if it's because it came at the end of the survey when we'd exhausted the patience of our poor sample population:

```{r}
long_bad_tc |>
  filter(name == nom[7]) |>
  slice_sample(n = 5) |>
  knitr::kable(caption = "5 random responses from a follow-up question about a second-favourite")
```

## Acknowledgments

Many thanks to the several hundred members of the KIND network who donated their time and brain-power to trying to complete deliberately badly-built surveys. I'm especially indebted to Kenneth Mack (Healthcare Improvement Scotland) who suggested that TV programme preferences would be a good venue to explore issues about building effective surveys.

::: {.content-visible when-format="revealjs"}
## Forthcoming sessions
```{r}
KINDR::training_sessions(start_date = params[["sl_date"]], n = 5, hide_area = T)
```

## Feedback

Please can you [spare us one minute of your time to give feedback](`r KINDR::feedback_url(params[["fn"]], date = params[["sl_date"]]) `)?
:::



