---
title: "Dates and times with lubridate"
date: 2026-07-06
execute: 
  echo: true
categories: [R, beginner]
params:
  sl_date: 2026-07-06
  fn: "Dates and times with lubridate"
  id: KR18
---

```{r}
#| echo: false
#| results: asis
#| fig-align: left
#| fig-height: 0.6
#| fig-width: 3

source(here::here("R/feed_block.R"))
# feed_block("Dates and times with lubridate")
feed_block(params$id)
```


## About this session

+ this is a beginner's introduction to working with dates and times with lubridate
+ focus on core parsing, get/set, and rounding functions
+ lots on dates, a bit of date-times, no times
  
:::{.callout-tip collapse=false appearance='default' icon=true}
## Session resources
+ [Lubridate cheatsheet](https://rstudio.github.io/cheatsheets/html/lubridate.html)
+ [R4DS 2e chapter on dates and times](https://r4ds.hadley.nz/datetimes)
+ [quick primer on ISO8601](https://en.wikipedia.org/wiki/ISO_8601)
+ there's also a bit of optional sample data which you can [download as a .csv](data/lubridate.csv)
:::

## Why lubridate?

+ makes some of the simple things easier
  + also makes some of the complicated things possible, even though we won't talk about them here
+ note that [lubridate](https://cran.r-project.org/web/packages/lubridate/index.html) is:
  + not the only way of dealing with dates
  + not always the best, but on balance the most consistent, and least quirky tools for dates

## Getting started

1. Open a new Rstudio project
1. In the console, run `install.packages("lubridate")`
1. Once installation has completed, create a new R script and save it
1. Add `library(lubridate)` to the start of the script

## Dates in R

+ effectively an integer count of **days** since 1970-01-01 (basically [Unix time](https://en.wikipedia.org/wiki/Unix_time))
+ that means we can convert integers to dates using `as_date()`:

```{r}
library(lubridate)
as_date(0) # showing the start of Unix time
as_date(20640) # date-of-writing
as_date(-1000) # unlike Excel, R is fine with dates before 1970
as_date(1:3) # vectorised
class(as_date(1000)) # the date class is in base-R, not lubridate
as.numeric(as_date(1000)) # converting back from date to numeric
```

## Date-times in R

-  effectively an integer too, but of **seconds** since 1970-01-01 00:00:00 UTC
- this is a pain-point when coming from Excel. Excel dates are days since 01-01-**1900**, and datetimes/times are decimal proportions of days (so `1.5` in Excel = 36 hours)


```{r}
as_datetime(0)
as_datetime(1783327643) # date-time-of-writing
as_date(-1000) # unlike Excel, R is fine with dates before 1970
as_datetime(1:3) # vectorised
class(as_datetime(1000)) # date-times are held using the POSIXct class, again from base-R
as.numeric(as_datetime(1000)) # converting back from date to numeric
as_datetime(2 ^ 31-1) # 32 bit signed int, which might cause a few problems
```

## Time zones

+ these date-times have an implicit time zone built-in (which is [UTC](https://en.wikipedia.org/wiki/Coordinated_Universal_Time))
+ that's potentially very important for time calculations (e.g. when calculating shifts when the clocks go back or similar)
+ time zones are [also part of base R](https://rdrr.io/r/base/timezones.html)
+ we can set the time zone explicitly:

```{r}
as_datetime(1783327643, tz = "GB")
as_datetime(1783327643, tz = "Europe/London") # equivalent: don't think there are specifically Scottish codes though
```

+ you can find your local time zone using `Sys.timezone()`, and see a list of all time zones (hundreds) using `OlsonNames()`

## Parsing dates

+ converting dates in your raw data into proper R dates is known as **parsing**
+ `as_date` in our examples above started with a number of days or seconds
+ if we have dates in our data, we'll need to calculate those numbers of days/seconds
+ it's extremely important. For example...

```{r}
# spot the deliberate mistake

dat <- readr::read_csv(here::here("r_training/data/lubridate.csv"))

dat |>
  dplyr::arrange(value) |>
  dplyr::slice(1:10) |>
  ggplot2::ggplot(ggplot2::aes(x = date, y = value, group = 1)) +
  ggplot2::geom_point() +
  ggplot2::geom_line()

typeof(dat$date)
```


+ make sure that dates in your data are stored as dates/date-times, not chr
+ `str()` / `glimpse()` are your friends here
+ most functions that accept dates (like ggplot) will mis-behave if you feed them date-shaped-words
    + e.g. alphabetically-ordered dates in the plot above
+ we also often want to be able to calculate with dates
+ yet this sort of parsing is a pain, because there are loads of inconsistent and ambiguous ways of representing dates

## Parsing functions

+ taking human-readable dates (and date times) and parsing them is where lubridate shines

```{r}

as_date("2026-07-06") # good for ISO8601-ish dates
dmy("06*07/2026")
ymd("2026__07__06")

dmy_hms("6/7/26 10:32:01")
```


```{r}
#| fold: true

# fixed the deliberate mistake
dat |>
  dplyr::mutate(date = dmy(date)) |>
  dplyr::arrange(value) |>
  ggplot2::ggplot(ggplot2::aes(x = date, y = value, group = 1)) +
  ggplot2::geom_point() +
  ggplot2::geom_line()

```


## Inconsistent date formats

```{r}
date_input <- c("17/5/24", "2024-05-17", "Friday 17th May 2024", "17*May*24", "5/17/2024" )

dplyr::coalesce(dmy(date_input),
               ymd(date_input),
               mdy(date_input))

parse_date_time(date_input, orders = c("dmy", "ymd", "mdy")) # similar, but note date times

```


## What's the time now?

```{r}
today()
date_decimal(2024.37534)
now()
now("Japan")

random_zone <- sample(OlsonNames(), 1)
cat(paste("The date-time in", random_zone, "is", now(sample(OlsonNames(), 1))))
```

## Get and set

+ as well as parsing text into proper dates, lubridate has many get/set functions for accessing/changing parts of dates
+ we'll show these off using `today()` to easily generate a date

### Years

```{r}
year(today())
leap_year(today())
quarter(today())
semester(today())
semester(today(), with_year = T)
```

### Months and weeks

```{r}
month(today())
month(today(), label = T)
week(today())
epiweek(today()) # special ways of counting weeks. See  https://en.wikipedia.org/wiki/ISO_week_date and https://www.cmmcp.org/mosquito-surveillance-data/pages/epi-week-calendars-2008-2024
```

### Days

```{r}
day(today())
wday(today())
wday(today(), label = T)
qday(today())
```

### Hour and minute

```{r}
hour(now())
minute(now())
am(now())
dst(now()) 
```

### Set

```{r}
update(now(), hour = 11, minute = 0, second = 0) # date to date-time
test_date <- dmy("05/06/23")
day(test_date)
day(test_date) <- 11
test_date
```

## Round

```{r}
floor_date(today(), unit = "week")
round_date(today(), unit = "week")
ceiling_date(today(), unit = "month")
rollback(today()) # last day of previous month
```

## Excel dates

+ Excel dates start at 1900 rather than 1970

```{r}
as_date(45429 - 25569) # dirty but effective for dates because it's more human read-able
as_date(45429, origin = "1899-12-30") # better for dates, but note the off-by-one error in 1900-format dates
```

## Excel date times

+ Excel's date times are stored as days and fractions of a day
+ in R, that makes them much more horrible than dates, because we think about date times in seconds, rather than days
+ do the offset first, to account for the different starting points of the different systems
+ days have `60 * 60 * 24 = 86400` seconds, so then multiply the decimal part by 86400

```{r}
as_datetime((46209.43 - 25569) * 86400) # beware the rounding errors though

```

