---
title: dbplyr and dtplyr
date: 2026-08-04
categories: [R, intermediate]
draft: true
params:
  sl_date: 2026-08-04
  fn: "dbplyr and dtplyr"
  id: KR46
execute:
  echo: true
  output: "markup"
---

```{r}
#| echo: false
#| results: asis
#| fig-align: left
#| fig-height: 0.6
#| fig-width: 3

source(here::here("R/feed_block.R"))
feed_block(params$id)
here::i_am("r_training/free_text_wrangling.qmd")
```

## About this session



::: {.callout-tip collapse="false" appearance="default" icon="true"}
## Session resources
+ you don't need anything apart from an R setup for this session with a few packages installed (see the setup chunk below)
:::

## Setup

Create a new project, install packages if needed, attach the stringr, tidytext, and dplyr packages, and create some simple data:

```{r}
# install.packages("dplyr")
# install.packages("dtplyr")
# install.packages("dbplyr")
# install.packages("palmerpenguins")
# install.packages("DBI")

library(dplyr)
library(dtplyr)
library(dbplyr)
library(palmerpenguins)

```

## What?

+ dplyr is great! Easy to read and write, nicely compatible with other R tools, with a chunky user-base and nice docs
+ dplyr is also a) slow and b) holds your data in memory
+ that means that when your data gets big (or complicated, or remote,...) dplyr becomes slow and annoying
+ the dream: write dplyr, but *execute something else* more capable of tackling the big stuff
+ both dtplyr and dbplyr are both flavours of *execute something else* while writing dplyr
  + [dtplyr](https://dtplyr.tidyverse.org/articles/translation.html) = [data.table](https://r-datatable.com/)
  + [dbplyr](https://dbplyr.tidyverse.org/) = [SQL](https://www.w3schools.com/sql/)

## dtplyr

Why not just write data.table directly? Largely because it's very unlike ordinary R code:

```{r}
library(data.table)

peng_dt <- data.table(penguins)
peng_dt[island == "Dream"] # filter
peng_dt[,.(bill_length_mm)] # select
peng_dt[,bill_length_mm] # pull or $
peng_dt[,.N, island] # count
```


It's actually a great system if you have the time to spend getting comfortable with it: really a very concise way of writing SQL-ish expressions. But, it's not easy to improvise, and it really does take a bit of time to get happy working in this way, especially if you've spent lots of time writing nice literate tidyverse code.

Enter dtplyr. That only really needs us to learn one new function:

```{r}
peng_dtp <- dtplyr::lazy_dt(penguins) # create a dtplyr object

peng_dtp |>
  as_tibble() # if you need to get your results back as a proper tibble

peng_dtp |>
  collect() # similarly, returns a tibble
```

Otherwise you can largely just start plugging in dplyr code:

```{r}
peng_dtp |>
  filter(island == "Dream")

peng_dtp |>
  select(bill_length_mm)

peng_dtp |>
  pull(bill_length_mm)

peng_dtp |>
  count(island)
```

There are also some non-dplyr bits of tidyverse to play with too - check the [reference](https://dtplyr.tidyverse.org/reference/index.html):

```{r}
peng_dtp |>
  # select() |>
  tidyr::pivot_longer(contains("mm"))
```


If you want to find out what's going on in ~~Narnia~~ data.table land, you can use `show_query()`

```{r}
peng_dtp |>
  count(island) |>
  show_query()
```

This is pretty interesting, and a delightfully lazy way to brush up on your data.table syntax:

```{r}
peng_dtp |>
  filter(island == "Dream") |>
  select(island, bill_length_mm) |>
  mutate(bill_length_rods = bill_length_mm / 5029.2) |>
  show_query()

```


```{r}
peng_dtp_inter_nc <- peng_dtp |>
  mutate(bill_vol = bill_length_mm * bill_depth_mm) 

peng_dtp_inter <- peng_dtp |>
  mutate(bill_vol = bill_length_mm * bill_depth_mm) |>
  compute() # intermediate assignment

waldo::compare(peng_dtp_inter_nc, peng_dtp_inter)
```

## dbplyr

dbplyr takes a similar approach to dtplyr, but has much more flexibility about source data. You can easily e.g. connect to a remote server or similar using DBI. That's largely outwith the scope of this session, so we'll create an in-memory SQLite database as a demonstration of how this might be used to manage large datasets in R:

```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), dbname = ":memory:") # creates a SQLite db

copy_to(con, penguins, "pengs") # adds the penguins dataset as a table called pengs
```

We could access this db directly using SQL queries:

```{r}
con |> 
  DBI::dbGetQuery('SELECT * FROM pengs WHERE island == "Dream" LIMIT 5') # filter to keep the preview a sensible length

con |> 
  DBI::dbGetQuery('SELECT bill_length_mm FROM pengs LIMIT 5') # select

con |>
  DBI::dbGetQuery('SELECT island, count(*) as n FROM pengs GROUP BY island')
```

Again, SQL is great if you have time to learn it properly, but for occasional use (or learning) dbplyr is even better:

```{r}
tbl(con, "pengs") |>
  filter(island == "Dream") # to return a db preview

tbl(con, "pengs") |>
  filter(island == "Dream") |>
  collect() # to return a proper tibble

tbl(con, "pengs") |>
  filter(island == "Dream") |>
  select(species) |>
  show_query() # for L&D purposes

tbl(con, "pengs") |>
  filter(island == "Dream") |>
  select(species) |>
  explain()

tbl(con, "pengs") |>
  filter(island == "Dream") |>
  select(species, bill_depth_mm) |>
  compute(name = "dream_pengs")

tbl(con, "dream_pengs") |>
  collect()

new_peng <- copy_inline(con, tibble(species = "Humboldt", bill_depth_mm = 31))

rows_insert(tbl(con, "dream_pengs"), new_peng, conflict = "ignore") |>
  count(species) |>
  compute(name = "species_pengs")

tbl(con, "species_pengs") |>
  collect()
```
