---
title: Exploring data with dplyr
date: 2025-05-19
execute: 
  echo: true
  output: "markup"
  freeze: auto
categories: [R, intermediate]
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| results: asis
#| fig-align: left
#| fig-height: 0.6
#| fig-width: 3

source(here::here("R/feed_block.R"))
feed_block("Exploring data with dplyr")
```


## Session outline

This session is an 🌶🌶 **intermediate practical** designed for those with some R experience. The aim of this session is to do three things with dplyr:

+ build your fluency with basic dplyr functionality
+ show some of the similarities of approach across the package, again to help you master the syntax
+ learn some of the more advanced functions to harness the power available beneath the surface

You might also like some of the other dplyr-themed practical sessions:

* [Summarising data with dplyr](dplyr_group_summarise.qmd)
* [Joining data with dplyr](dplyr_joins.qmd)
* [Tidyselect](tidyselect.qmd)

## A starting-point

You should have at least some previous experience working with dplyr before starting this training. Specifically, you should be familiar with at least:

+ `select()`, to select columns from data
+ `filter()`, to select rows from data
+ `mutate()`, to make new columns from existing columns

## Investigating data

Dplyr is usually used as a toolkit for manipulating tidy data. This session concentrates on some lesser-used functions in the dplyr package that are especially useful earlier in the analysis lifecycle when exploring new data: `slice()`, `glimpse()`, `rename()`, `relocate()`, `case_when()`/`case_match()`, and `consecutive_id()`

Package loading as follows:

```{r}
library(NHSRdatasets)
library(dplyr)
```


### `slice()`

It's definitely worth pointing you towards the [slice() manual page](https://dplyr.tidyverse.org/reference/slice.html) - it's clear and interesting.

At its simplest, `slice()` is an alternative to head:
```{r }
ae_attendances |> 
  slice(1:6)
```

The number supplied to `slice()` returns that row - so `slice(3)` shows the third row. You can also provide a sequence or vector of rows:

```{r }
ae_attendances |> 
  slice(1:3)

ae_attendances |> 
  slice(2,5,918)
```

More interestingly, you can `group_by()`, then slice, to see the first row of each group. Here, the data is filtered, grouped by org_code, then we slice the top row of each group:

```{r}
ae_attendances |>
  filter(attendances >= 22000) |>
  group_by(org_code) |>
  slice(1)
```

(we'll explore `group_by()` in the [summarising data with dplyr](dplyr_group_summarise.qmd) session, so don't worry if this is unfamiliar)

The `slice_min()` and `slice_max()` functions are also useful, giving you something like an integrated `arrange()` for looking at slices of the data:

```{r}
ae_attendances |>
  filter(type == "1") |>
  slice_max(breaches, n=1)


# equivalent to
ae_attendances |>
  arrange(desc(breaches)) |>
  slice(1)

```
(note that you need to explicitly name the number of rows you want using `n=...`)

`slice_min()` and `slice_max()` are particularly useful for groups:

```{r}
ae_attendances |>
  filter(type == "1") |>
  group_by(org_code) |>
  slice_max(breaches, n=1) 
```

`slice_sample()` gives a random sampling of the data.
```{r}
ae_attendances |>
  slice_sample(n=6) 
```

There are also `slice_head()` and `slice_tail()` functions which work in the same way.

Final tip - slice with a negative index can be used as like `filter()` to remove the specified row(s):
```{r}
# returns the last 5 rows only
ae_attendances |> 
  slice(-1:-12760)
```

### `glimpse()`

`glimpse()` is a data viewing function similar to `print()`, except it transposes the data so that each column is displayed as a row. This is particularly useful for wide data with many columns, especially when you are interested in checking the class (date, character, etc.) of your columns.

```{r}
synthetic_news_data |> 
  glimpse() 
```

Note the useful dimension information in the first two lines, and the class information in angle brackets. Okay, so there are lots of similar ways of displaying the same information, but `glimpse()` is nice and concise both to write and to read. Compare a couple of base r (or purrr) near-equivalents, which are especially messy when the data is quite wide:

```{r}
synthetic_news_data |> 
  purrr::map_df(class)

# equivalent in base R
lapply(synthetic_news_data, class)
```

`glimpse()` plays nicely with the pipe, meaning that it's potentially useful while you're working on a complicated data transformation. If you have lots of stages piped together, you can insert `glimpse()` in periodically to check that each stage of your transformation is working as expected:

```{r}
synthetic_news_data |> 
  glimpse() |>
  filter(age == 71 & male == 0) |>
  glimpse() |>
  mutate(pulse_pres = syst-dias) |>
  glimpse()

```


### `rename()` (and `rename_with()`)

`rename()` renames columns. Like `mutate()`, the general syntax for `rename()` is *new name* = *old name*.

```{r}
LOS_model |>
  rename(age = Age) 
```

There's not much more to say about the basic `rename()` function beyond that. However, the scoped variants, `rename_with()` and `rename_all()` provide some useful additional tools. For example, we can use `tolower()` - which converts strings to all lower-case - to rename multiple columns at once:

```{r}
LOS_model |> 
  rename_with(tolower) 
```
`rename()` works well with [tidyselect operators](tidyselect.qmd) too, as we discuss in that session.

### `relocate()`

`relocate()` re-arranges the order of columns:

```{r}
LOS_model 
```

`relocate()`'s default behaviour is to move specified columns to the far left:
```{r}
LOS_model |>
  relocate(LOS) 
```

You can specify `.before` and `.after` if you need finer control over where your column ends up:

```{r}
LOS_model |>
  relocate(LOS, .after=Death)

LOS_model |>
  relocate(Death, .before=last_col()) # little bit of tidyselect
```

### `case_when()`

The ONS mortality data in NHSRdatasets has several categories
```{r}
ons_mortality |>
  count(category_1)
```

Winter (metrological winter, for clarity) runs from the start of December to the end of February. Can we class the total deaths into winter, and non-winter categories? In practice, I'd be tempted to use `lubridate::quarter()` with an appropriate starting offset for this kind of work if possible, especially if I was grouping into regular groups. Winter/non-winter isn't regular though, so `case_when()` is the better option. Here, we'll work in stages. First, as we'll need the months, we'll extract those using `lubridate::month()`:

```{r}
ons_mortality |>
  filter(category_1 == "Total deaths") |>
  mutate(month = lubridate::month(date))
```

Our definition of winter is any case where that month is 1, 2, or 12. So use `case_when()` to catch Jan and Feb:

```{r}
ons_mortality |>
  filter(category_1 == "Total deaths") |>
  mutate(month = lubridate::month(date)) |>
  mutate(thing = case_when(month < 3 ~ "Winter"))
```

and then add an extra case to deal with December:

```{r}
ons_mortality |>
  filter(category_1 == "Total deaths") |>
  mutate(month = lubridate::month(date)) |>
  mutate(season = case_when(month < 3 ~ "Winter",
                           month == 12 ~ "Winter"))
```

And finally a default to catch all the non-Winter values:

```{r}
seasonal_ons_mortality <- ons_mortality |>
  filter(category_1 == "Total deaths") |>
  mutate(month = lubridate::month(date)) |>
  mutate(season = case_when(month < 3 ~ "Winter",
                           month == 12 ~ "Winter",
                           TRUE ~ "Not winter"))
seasonal_ons_mortality

seasonal_ons_mortality |>
  count(season)
```

### `case_match()`

Rather than a logical test, `case_match()` matches against values rather than performing a logical test. This is excellent for the winter classification example:

```{r}
ons_mortality |>
  filter(category_1 == "Total deaths") |>
  mutate(month = lubridate::month(date)) |>
  mutate(season = case_match(month, c(1, 2, 12) ~ "Winter",
                             .default = "Not winter"))
```

### `consecutive_id()`

```{r}
consecutive_id(c(1, 1, 1, 2, 1, 1, 2, 2)) # puts groups together by changes. This would have been very useful for my transcripts

repeats <- tibble(speaker = c("steve", "steve", "emma", "steve", "emma", "emma"),
       comment = letters[1:6]) # repeated data

repeats |>
  mutate(group = consecutive_id(speaker)) # add a group that increments with each change of speaker

repeats |>
  mutate(group = consecutive_id(speaker)) |>
  group_by(group) |>
  mutate(comment = paste(comment, collapse = " ")) |>
  distinct()
```



