---
title: "Working with lists in purrr"
date: 2026-08-20
execute: 
  echo: true
  output: "markup"
  draft: true
categories: [R, intermediate]
params:
  sl_date: 2026-08-20
  fn: "Working with lists in purrr"
  id: KR64
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| results: asis
#| fig-align: left
#| fig-height: 0.6
#| fig-width: 3

source(here::here("R/feed_block.R"))
# feed_block("Working with lists in purrr")
# feed_block(params$id)
```

## Introduction

purrr is mainly used for functional programme, especially making it easier and more consistent to apply a function over a group of items and collect the output:

```{r}
more_less <- function(col) {
  if(sum(col) > 200) "Loads"
  else "Not much"
}

library(purrr)

mtcars |>
  map(more_less)

mtcars |>
  map_vec(more_less)
```

But purrr also has a set of tools for working with lists. This session is an introduction to those functions. We'll use the [purrr cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/purrr.pdf) as a starting point for this session.

## A reminder about lists

Lists are a data structure in R with two very useful properties. Unlike vectors, they can contain data of different classes. Unlike data frames/tibbles, they can be ragged, containing items of different lengths:

```{r}
test_list <- list(a = LETTERS, b = NA, c = 1:5, d = NULL, e =  letters)

test_list
```

* subset a list by item/index with `[]`:

```{r}
test_list["c"]
test_list[1]
```

** retrieve the contents of a list item with `$` or `[[]]`:

```{r}
test_list[["c"]]
test_list$a
```

** subset those list contents with chained square brackets:

```{r}
test_list[["a"]][2]
test_list$c[3]
```




## `modify()`

Effectively `map` for list items. `modify()` applies a function to each list item and returns a structurally-identical modified list:

```{r}
test_list |>
  modify(toupper)
```

Select list items to modify with `modify_at()`/`modify_if()` / `modify_depth()`:

```{r}
test_list |>
  modify_at("e", toupper) # select list item by quoted name

test_list |>
  modify_if(is.numeric, \(x) x + 2) # select list item by function

list(list(list(list(deepo = c(3, 4, 5))))) |> # likely to be more trouble than it's worth, especially because many functions will do odd things to list items
  modify_depth(4, \(x) x * 10)
```

## `reduce()`

A bit like `cumsum()`! Apply a function to each element recursively.

```{r}
1:10 |>
  reduce(sum) # so think 1+2=3, then 3+3=6, then 6+4=10,...,45+10=55. That can be a bit hard to understand, so accumulate helps by showing intermediate results

1:10 |>
  accumulate(sum)

list(a = c("space", "star", "dragon"), "fruit") |>
  reduce(paste0) # odd results over vectors, and beware of vector recycling

```

## Vectors

Some purrr functions are designed for vectors:

```{r}
c(NULL, "hello", NULL, "world") |>
  compact()

c("NULL", "hello", "NULL", "world") |>
  keep_at(c(2,4))

c("NULL", "hello", "NULL", "world") |>
  discard_at(c(1,3))

c("NULL", "hello", "NULL", "world") |>
  set_names(c("one", "two", "three", "four"))

```

## Predicate functions

```{r}

test_list |>
  keep(is.numeric)

test_list |>
  discard(is.character)

test_list |>
  head_while(is.character)

test_list |>
  tail_while(is.character)

test_list |>
  detect(is.numeric)

test_list |>
  detect_index(is.numeric)

test_list |>
  every(is.character)

test_list |>
  some(is.character)

test_list |>
  none(is.complex)

test_list |>
  has_element(1:5)
```


## Pluck

```{r}
test_list |>
  pluck("e")

test_list |>
  pluck("f")

try({
  test_list |>
    chuck("f") # more surly than pluck
  })

test_list |>
  pluck_depth()

test_list |>
  assign_in("d", "boo!")

test_list |>
  modify_in("c", sqrt)
```

## Reshape and concatenate
```{r}
test_list |>
  list_flatten() # doesn't do anything: designed to flatten a list back to depth = 1

list("hello", "world") |>
  list_c() # strict about types, wouldn't try and combine our test_list

list(list(a = letters, b = LETTERS), list(c = letters, d = LETTERS)) |>
  transpose()

list(1, list(2), list(list(3))) |>
  list_flatten() |> # removes one level each use
  list_flatten()

```

