---
title: "The purrr package in R"
date: "2024-07-29"
execute: 
  echo: true
  eval: true
  freeze: auto
output: "markup"
categories: [R, intermediate]
---

::: {.callout-note collapse="false" appearance="default" icon="true"}
## Session materials

-   [all materials {{< iconify ph:file-zip size=2x >}}](src/purrr_package.zip)
-   slides [{{< iconify ph:file-html size=2x >}} html](src/purrr_package.html) / [{{< iconify ph:file-pdf size=2x >}} pdf](src/purrr_package.pdf)
:::


```{r}
#| echo: false
#| results: asis
#| fig-align: left
#| fig-height: 0.6
#| fig-width: 3

source(here::here("R/feed_block.R"))
feed_block("The purrr package in R")

```


## Welcome

-   this session is for 🌶🌶 intermediate users
-   you'll need R + Rstudio / Posit Workbench / posit.cloud to follow along

## Session outline

-   a digression about Linnaeus
-   functionals
-   base-R functional programming
-   `map` and `walk`
-   `map2` and `pmap`
-   niceties and add-ons

## A digression about Linnaeus

::: columns
::: {.column width="40%"}
![[1758 edition of the Systema Naturae](https://www.biodiversitylibrary.org/item/10277)](images/clipboard-4148155250.png){height="400"}


:::

::: {.column width="60%"}
+ used an existing system of binomial classification
+ *Homo sapiens*
    + *Homo* = generic name, which applies to similar species
    + *sapiens* = specific name, for that species and that species only
+ *Pan troglodytes* and *Pan paniscus* = two similar species in a genus
+ *Elephas maximus* and *Loxodonta africana* = two similar species in different genera



:::

:::

## Functionals

Here are some numbers:

```{r}
n1 <- 7:9
```



Let's find their average. We'd usually do this by passing those numbers to a function:

```{r}
mean(n1)
```

But in R, interestingly, we can also do this the other way round by passing a function name:

```{r}
my_num_f <- function(funct = mean) funct(n1)
my_num_f(mean)
my_num_f(sum)
```

We'd describe this as a [functional](https://adv-r.hadley.nz/functionals.html). It's fun, but a bit messy and annoying (e.g. how to change the numbers you're averaging??).

## Functional programming in base R

Say we've got a function we want to apply:

```{r}
round_root <- function(n) round(n ^ 0.5, 1)
```

There are several ways of applying functions to stuff in base R. 
+ we could use a loop: that's another session
+ we could just exploit the vectorised nature of most functions in R
```{r}
round_root(n1)
```

+ or we could use some of the `apply` family of functions, like `lapply` and `sapply`:

```{r}
lapply(n1, round_root) # returns a list
sapply(n1, round_root) # simplifies that list to a vector

```

There's no real reason to use these functions when things are this simple, but when our applications become more complicated...

```{r}
n2 <- 11:13

lapply(list(n1, n2), round_root)
sapply(list(n1, n2), round_root) # oddball output
lapply(list(n1, n2[1:2]), round_root) # quirky

```

## purrr

+ [purrr](https://cran.r-project.org/web/packages/purrr/index.html) is a functional programming toolkit
+ main advantage = **consistency**
+ [very useful cheatsheet](https://rstudio.github.io/cheatsheets/html/purrr.html)

## `map`

`map` is our purrr [type specimen](https://en.wikipedia.org/wiki/Type_(biology))

```{r}
library(purrr)
map(n1, round_root)
```

Pleasingly, `map` will handle all kinds of odd inputs without fuss:

```{r}
map(c(n1, n2), round_root)
map(dplyr::tibble(n1 = n1, n2 = n2), round_root)
map(rbind(n1, n2), round_root) 
```

`map` will always return a list - that's because, no matter what the output, you can always cram it into a list. If you want different output, you can have it. You just need to find the right *species*:

```{r}
map_vec(n1, round_root)
try(map_int(n1, round_root)) # surly and strict
```

```{r}
round_root_int <- function(n) as.integer(n ^ 0.5)
map_int(n1, round_root_int) 
```

```{r}
round_root_lgl <- function(n) as.integer(n ^ 0.5) %% 2 == 0
map_lgl(n1, round_root_lgl)
```

## anonymous functions

If you're comfortable with the new anonymous function syntax, you can build an anonymous function in place:

```{r}
map_lgl(1:4, \(x) x %% 2 == 0)
```


## `walk`

`walk` is intended for code where the side-effect is the point: graphs, pipes, and Rmarkdown especially. Otherwise, it's as `map`:

```{r}
walk(n1, round_root) # wtf?
round_root_print <- function(n) print(n ^ 0.5)
walk(n1, round_root_print) 
```

```{r}
round_root_cat <- function(n) cat(n ^ 0.5, "  \n")
walk(n1, round_root_cat)
```


## `map2` 

`map2` is for 2-argument functions:

```{r}
map2_int(n1, n2, `+`) # the best terrible way of adding I know
round_root_places <- function(n, dp = 1) round(n ^ 0.5, dp)
round_root_places(n1, 0)
map2(n1, 0, round_root_places)
```

## Beware of recycling rules

You'll be unable to use `map2` if your inputs are different lengths:

```{r}
try(map2(1:3, 0:3, round_root_places))
```

This makes `expand.grid` valuable if you're looking to try out all the combinations of two vectors, for example.

```{r}
dat <- expand.grid(nums = 1:3, dplaces = 0:3) 

map2(dat$nums, dat$dplaces, round_root_places)
```

Or (probably more usefully) this could be done inside `dplyr::mutate` to add to a tibble:

```{r}
#| results: asis
expand.grid(nums = n1, dplaces = 0:3) |>
  dplyr::as_tibble() |>
  dplyr::mutate(rr = map2_vec(nums, dplaces, round_root_places)) |>
  knitr::kable()
```

## `pmap`

`pmap` is for n argument functions.

```{r}
round_roots_places <- function(n, root = 2, places = 1) round(n ^ 1/root, places)

round_roots_places(n1, root = 4, places = 2) # use named arguments to avoid misery

pmap(list(n = n1, root = 4, places = 2), round_roots_places)

```

## Niceties and addons

`imap` allows you to work with indicies, and `list_c` converts simple lists back to vectors:

```{r}
imap(list("a", "b", "c"), \(x, y) paste0(y, ": ", x)) |> # index map where y is the name or index
  list_c()
```

If your function returns a tibble, you can use `list_rbind` or `list_cbind` to row- or column-bind the results into a tibble:

```{r}
#| results: asis
map(n1, \(x) dplyr::tibble("Val" = x, "sq_val" = x^2)) |>
  list_rbind() |>
  knitr::kable()
```

## Resources

- [functionals chapter in Advanced R](https://adv-r.hadley.nz/functionals.html)
- [purrr cheatsheet](https://rstudio.github.io/cheatsheets/html/purrr.html)
