---
title: An introduction to stringr
date: 2026-07-28
draft: true
categories: [R, intermediate]
params:
  sl_date: 2026-07-28
  fn: "An introduction to stringr"
  id: KR61
execute:
  echo: true
---

```{r}
#| echo: false
#| results: asis
#| fig-align: left
#| fig-height: 0.6
#| fig-width: 3

source(here::here("R/feed_block.R"))
feed_block(params$id)
```

## About this session

This is a stand-alone beginner/intermediate introduction to using the stringr package in R to manipulate text. It's pitched somewhere between a complete introduction and a more intermediate skill-builder session, so somewhere between our usual beginner (🌶) and intermediate (🌶🌶) levels. In this session:

+ we'll introduce stringr and talk about what it's for
+ we'll look at a core function toolkit
+ and we'll compare and contrast to some other ways of working with text in R
+ we **won't** try and spend a lot of time thinking about other aspects of text wrangling, or using regular expressions here as we offer other sessions designed to cover that ground

::: {.callout-tip collapse="false" appearance="default" icon="true"}
## Session resources
- [a set of words generated by blended LLM and KIND member labour](data/llm_words.xlsx)
:::

```{r}
library(dplyr)
library(stringr)

words <- readxl::read_xlsx("data/llm_words.xlsx") 

words |>
  head() |>
  knitr::kable()
```

