---
title: Stringdist
date: 2024-09-23
execute: 
  echo: true
  eval: true
  output: true
  freeze: auto
categories: [R, intermediate, NLP]
editor_options:
  chunk_output_type: console
params:
  sl_date: 2024-09-23
  fn: "Stringdist"
  # id: KP02
---

```{r}
#| echo: false
#| results: asis
#| fig-align: left
#| fig-height: 0.6
#| fig-width: 3

source(here::here("R/feed_block.R"))
# feed_block(params$id)
#feed_block("stringdist")
```


## Introduction
This is an introduction to [stringdist](https://cran.r-project.org/web/packages/stringdist/stringdist.pdf) which is designed to allow fuzzy matching of text. Install and attach in the standard way:

```{r}
# install.packages("stringdist") # likely needs compiling
library(stringdist) 
library(ggplot2) # for something fun later
```

While there are several families of functions in the package, we'll concentrate on those that focus on the core idea: strings differ, and you can quantify those differences for pleasure and profit. A simple example:

```{r}
stringdist("thing", "think")
stringdist("thing", "string")
stringdist("thing", "oxpecker")
```

Once you've got an idea about how different you're okay with your strings being, you can use e.g. a max distance as a cut-off to do work in your code. A simple example of implementing this can be found in the `amatch` function, which is designed to work like a fuzzy matching version of base-R's `match`. We'll use `a` and `b` as variable names to keep aligned with the stringdist documentation. Quick refresher on `match`:

```{r}
a <- "the" # take a value
b <- c("this", "is", "the", "thing") # and a vector of possible values
match(a, b) # match returns the index location of your a in your b
```
Match lets you do useful work with strings - e.g. remove everything before your key word:

```{r}
b[match(a, b):length(b)]
```

## exemplar: `amatch`

And now for the `amatch` version:

```{r}
a <- "thee" # take a value
amatch(a, b, maxDist = 1) # maxDist = allowable number of typos
amatch(a, b, maxDist = 10) # where several possible values exist, amatch returns the closest match
```

You can use that in a base-R way to e.g. correct wonky input values:
```{r}
b[amatch(a, b, maxDist = 1)] 
```

Or you might prefer the ain helper:
```{r}
ain(a, b, maxDist = 1)
a %in% b # like a fuzzy version of %in%
"the" %in% b
```

Or in a slightly horrid but tidyish way if you wanted to tidy up several different values:
```{r}
dplyr::tibble(raw_data = c("thi", "ss", "thee", "thung")) |>
  dplyr::rowwise() |>
  dplyr::mutate(is_match = b[amatch(raw_data, b, maxDist = 5)])
```

## other functions

### `stringdist`, `stringsim`, and `stringdistmatrix`
`stringdist` has a nice toolkit of related functions. `stringdist` gives you the distance between strings:

```{r}
stringdist("thing", "think") # gives you the raw distance.
stringdist(a, b) # vectorised
```

`stringsim` does the opposite, giving a similarity score between 0 and 1:

```{r}
stringsim(a, b)
stringsim("cat", "dog")
stringsim("cat", "cat")
```

`stringdistmatrix` does what you might imagine, returning a matrix of differences. This is largely useful if you want e.g. to pass these scores into something like a clustering algorithm that expects a matrix:

```{r}
stringdistmatrix(a, b) # same idea, different output: a matrix of differences
hclust(stringdistmatrix(b))
plot(hclust(stringdistmatrix(b, useNames = "strings")))
```

It's possible-but-messy to take this matrix and tidy it into a tibble:
```{r}
stringdistmatrix(c("thee", "che"), b, useNames = "strings") |>
  as.data.frame() |>
  tibble::as_tibble(rownames = "a") |> # slightly roundabout conversion to tibble via df to preserve rownames
  tidyr::pivot_longer(!a, names_to = "b", values_to = "stringdist") # useful 
```

But probably better/more idiomatic to go:
```{r}
tidyr::expand_grid(a = c("thee", "che"), b) |>
  dplyr::mutate(stringdist = stringdist(a, b))
```

### `afind`, `grab`, `grabl`, and `extract`

```{r}
afind(a, b)
```

`afind` returns a lot of stuff:

```{r}
afind(a, b)$location # where in substrings do the search terms start?
afind(a, b)$distance # how far from the search term is each substring?
afind(a, b)$match # what's the best matching part of the search term against each substring
```

If that's a bit confusing, `grab` and `grabl` are basically replacements for `grep(l)`. 
`grab` == `grep` and `grabl` == `grepl`:

```{r}
c <- c("wig", "win", "banana", "ring")
c[grab(c,  "wig", maxDist = 1)]
c[grab(c,  "wig", maxDist = 2)]
grabl(c,  "wig", maxDist = 2)
c[grab(c,  "wig", maxDist = 3)]
extract(c, "wii", maxDist=1)
```

## What do you mean closest?

The *really* interesting bit: you can play with the method used to calculate distance which will give different results:

### hamming
Hamming distance. Number of character substitutions required to change a into b.

```{r}
amatch("thign", b, maxDist = 1, method = "hamming") # not found
amatch("thign", b, maxDist = 2, method = "hamming") # n subs g, g subs n = 2
```

### lv
Levenshtein distance. Number of del/ins/subs required to change a into b.

```{r}
amatch("thinng", b, maxDist = 2, method = "hamming")
amatch("thinng", b, maxDist = 2, method = "lv") # as we can now delete
```

### osa
Optimal string aligment. Number of del/ins/subs/swaps required to change a into b:

```{r}
amatch("think", b, maxDist = 1)
amatch("thign", b, maxDist = 1, method = "osa") # n swaps with g
```

### dl
Full Damerau-Levenshtein distance. As osa, but calculated slightly differently.

```{r}
amatch("thign", b, maxDist = 1, method = "dl") # can't find a difference with these simple ones from osa
```

### lcs
Longest common substring distance. Comparing longest identical parts of a and b:
```{r}
amatch("thign", b, maxDist = 2, method = "lcs") # I think roughly = length - common substring
```

### soundex
Distance based on soundex encoding. Used in the `phonetic` function (see below), but finds homophones:

```{r}
amatch("thign", b, maxDist = 1, method = "soundex")
stringdist("too", c("two", "to", "oto"), method = "soundex")
```

There are also several methods designed to look at **q-grams** = substring chunks of q characters long, and a group of methods based on the number of similar and different q-grams between a and b:

### qgram
This is the q-gram distance:
```{r}
stringdist("thign", b, method = "qgram")
```

### cosine

The cosine distance between q-gram profiles, which gives you (usefully) a scaled distance of between 0 (identical) and 1 (utterly different)

```{r}
stringdist("thing", b, method = "cosine")
stringdist("thing", "fox", method = "cosine")
```

### jaccard

The [Jaccard distance](https://en.wikipedia.org/wiki/Jaccard_index) between q-gram profiles

```{r}
stringdist("thing", b, method = "cosine") 
stringdist("thing", b, method = "jaccard") 
```

### jw

The Jaro distance (and Jaro-Winkler distance):

```{r}
stringdist("thing", b, method = "jw", p = 0.2) # correction term
```

## phonetic

There's also tooling to deal with the similarity of phonetic spellings. Excellent for e.g. Teams transcripts, if you're trying to tidy those up:

```{r}
phonetic("ring")
phonetic("wing") # similar sounding strings get similar values
phonetic("string")
phonetic("strung")
```


## a play example: GGC

The KIND network has lots of members from the main NHS body in Glasgow, which is officially known as ["NHS Greater Glasgow and Clyde"](https://www.nhsggc.scot/). That's a highly variable and abbreviatable name though, leading to over 70 different permutations in common use. Here are some real-ish examples

```{r}
ggc <- readr::read_lines(here::here("r_training/data/ggc.txt")) # real data

sample(ggc, 10)
```

There's lots here to play with using `stringdist`. My two quick examples are to plot a couple of different sensible variants of the name ("NHS GGC", "NHS Greater Glasgow and Clyde", "NHSGGC") methods, and see how many distant results we get:

```{r}
tidyr::expand_grid(a = c("NHS GGC", "NHS Greater Glasgow and Clyde", "NHSGGC"), b = ggc) |>
  dplyr::mutate(stringdist = stringdist(a, b)) |>
  ggplot() +
  geom_density(aes(x = stringdist, fill = a)) +
  facet_wrap(~a, ncol = 1) +
  theme(legend.position = "none")
```

Or to try a family tree (more swishily, a [cluster dendrogram](https://www.statisticshowto.com/hierarchical-clustering/)) of those terms:

```{r}
sort(table(ggc), decreasing = T)[1:7] |>
  stringdistmatrix(useNames = "names") |>
  hclust() |>
  plot()
```
