---
title: An introduction to stringr
date: 2026-07-28
draft: true
categories: [R, intermediate]
params:
  sl_date: 2026-07-28
  fn: "An introduction to stringr"
  id: KR61
execute:
  echo: true
  eval: false
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| results: asis
#| fig-align: left
#| fig-height: 0.6
#| fig-width: 3

source(here::here("R/feed_block.R"))
feed_block(params$id)
```

## About this session

This is a stand-alone beginner/intermediate introduction to using the stringr package in R to manipulate text. It's pitched somewhere between a complete introduction and a more intermediate skill-builder session, so somewhere between our usual beginner (🌶) and intermediate (🌶🌶) levels. In this session:

+ we'll introduce stringr and talk about what it's for
+ we'll look at a core function toolkit
+ and we'll compare and contrast to some other ways of working with text in R
+ we **won't** try and spend a lot of time thinking about other aspects of text wrangling, or using regular expressions here as we offer other sessions designed to cover that ground

::: {.callout-tip collapse="false" appearance="default" icon="true"}
## Session resources
- [a set of words generated by blended LLM and KIND member labour](data/llm_words.xlsx), which you should download into a new Rstudio/posit project
+ you'll definitely want the [stringr cheatsheet on hand](https://rstudio.github.io/cheatsheets/strings.pdf)
:::

## Setup

Create a new project, attach the stringr and dplyr packages, load the data, and inspect:

```{r}
#| echo: false
#| eval: true
library(dplyr)
library(stringr)

words <- readxl::read_xlsx(here::here("r_training/data/llm_words.xlsx"))
```

```{r}
#| eval: false
library(dplyr)
library(stringr)

words <- readxl::read_xlsx("data/llm_words.xlsx") 
```

```{r}
#| echo: false
#| eval: true
words |>
  head() |>
  knitr::kable()
```

We'll also create a nice short vector of words to help us get up and running with our first few stringr functions:

```{r}
wd <- words$Human1[1:6]
```


## Detect

```{r}
str_detect(wd, "No") # logical output, vectorised, tells you which item contains the pattern

wd |>
  str_detect("No") # pipe-able, probably best when things get more complicated

str_which(wd, "No") # index, vectorised

str_detect(wd, "no") # case sensitive

str_detect(wd, "(?i)no") # horrid but handy

str_detect(wd, "N") # supply string or substring from anywhere in the string

str_starts(wd, "N") # find strings start that start with the pattern

str_starts(wd, "N", negate = T) # finds strings that do not start with the pattern

str_ends(wd, "o|e") # find at end, handy shorthand for several options

str_detect(wd, ".") # the pattern is really a regex - separate session!

str_count(wd, "o") # counts occurrences

str_locate(wd, "o") # note class, and beware: how many times does o occur in "chocolate"?
```

## The naming of things

+ lovely hierarchical naming convention: everything begins `str_`, and varient functions are usually named using the `_all` suffix:

```{r}
str_locate_all(wd, "o") # note class again
```

## A puzzle for you

Given what you already know about stringr, can you write some code to extract all and only the o-containing words from the `wd` vector please? You should return that as a new vector, containing just the o-words.

```{r}
wd[str_detect(wd, "o")]
wd[str_which(wd, "o")]
wd[str_detect(wd, "(?i)o")] # for the suspicious
```

## Subsetting

stringr gives you a shorthand way of doing that o-finding work:

```{r}
str_subset(wd, "o") # that is, it's subsetting your original vector

str_extract(wd, "o") # not very useful with the single letter

str_extract(wd, "^[A-M].*") # much more useful with regular expressions

str_extract_all(wd, "(?i)[l-n][aeiou]") # bit head-bending

str_view(wd, "(?i)[l-n][aeiou]") # might be helpful to know about

str_sub(wd, start = 1, end = 3) # by character index

str_sub(wd, start = -2) 
```

## Sorting

```{r}
str_sort(wd) # that returns a sorted vector
str_order(wd) # gets you the indices
```

## Lengths

```{r}
str_length(wd) # same as nchar(wd)

str_trunc(wd, 6)

str_trunc(wd, 6, ellipsis = "")

str_pad(wd, 9, pad = "_", side = "right") 

c("123456789", "1987654321") |> # possibly grubby way of fixing broken CHIs?
  str_pad(10, pad = "0")

c(" space before", "space afterwards ", "double  space") |>
  str_trim()

c(" space before", "space afterwards ", "double  space") |>
  str_squish()
```

## Change

```{r}
#| output: asis
str_to_lower(wd)
str_to_upper(wd)
str_to_title(wd)

str_replace(wd, "o", "_")
str_replace_all(wd, "o", "_")

str_c(wd, " is ", wd) # beware lengths of vectors
str_flatten(wd, " and ")

str_glue("Nine squared is {9*9}") # basically saves you loading the glue package

words |>
  slice_sample(n = 4) |>
  str_glue_data("{LLM1} and {LLM2} are some pairs of LLM words  \n")

str_split(wd, "o") # splits at each pattern and removes it, returning a list
str_split_i(wd, "o", i = 2) # split at pattern and return ith match
str_split_fixed(wd, "o", n = 2) # splits at each match, required n as a max value. Care with the class, and examine row 5

```

There's also a weird-but-useful way of modifying substring matches. Note this will alter your `wd` vector!

```{r}
str_sub(wd, start = 1, end = 3) <- "clonk" # assign into substituted string
wd
wd <- words$Human1[1:6] # recreate our original vector
```



