---
title: "Testing R code"
date: "2024-08-07"
execute: 
  echo: true
  eval: true
  freeze: auto
output: "markup"
categories: [R, intermediate]
---

::: {.callout-note collapse="false" appearance="default" icon="true"}
## Session materials

-   [all materials {{< iconify ph:file-zip size=2x >}}](src/testing.zip)
-   slides [{{< iconify ph:file-html size=2x >}} html](src/testing.html) / [{{< iconify ph:file-pdf size=2x >}} pdf](src/testing.pdf)
:::


```{r}
#| echo: false
#| results: asis
#| fig-align: left
#| fig-height: 0.6
#| fig-width: 3

source(here::here("R/feed_block.R"))
feed_block("Testing R code")
```


## Welcome

-   this session is for 🌶🌶 intermediate users
-   you'll need R + Rstudio / Posit Workbench / posit.cloud to follow along

## Session outline

+ introduction: why test?
+ informal testing
+ unit testing
    + introduction - why automate your tests?
    + `testthat` walkthrough

## Introduction: why test?

+ code goes wrong       
    + functions change
    + data sources change
    + usage changes
+ testing guards against the commonest problems
+ that makes for more reliable code
    + more reliable code opens the way to nicer development patterns

## A note

+ most discussions about testing come about as part of package development 
    + we'll avoid that area here, but please see the [three excellent chapters in the R packages book](https://r-pkgs.org/testing-basics.html)  for guidance
    + we'll also steer clear of Shiny/Rmarkdown/Quarto, as things can be a bit more tricky to test there
+ we also won't talk about debugging here (although do look out for the future training session on that)

## Informal testing

+ a real-world example: Teams transcripts
+ Teams transcripts can be very useful data-sources
+ but they're absolutely horrible to work with:

```{embed, file = "data/input.txt"}
```

+ imagine that you've written a (horrible) Teams transcript parser:
+ how would you test this code to make sure it behaves itself?

```{r}
#| echo: true

file <- "data/input.txt"

readLines(file) |>
    tibble::as_tibble() |>
    dplyr::filter(!value == "") |>
    dplyr::filter(!value == "WEBVTT") |>
    dplyr::mutate(question = stringr::str_extract(value, "^(Q.*?)::>$")) |>
    tidyr::fill(question, .direction = 'down') |>
    dplyr::filter(!stringr::str_detect(value,  "^(Q.*?)::>$")) |>
    dplyr::mutate(ind = rep(c(1, 2),length.out = dplyr::n())) |>
    dplyr::group_by(ind) |>
    dplyr::mutate(id = dplyr::row_number()) |>
    tidyr::spread(ind, value) |>
    dplyr::select(-id) |>
    tidyr::separate("1", c("start_time", "end_time"), " --> ") |>
    tidyr::separate("2", c("name", "comment"), ">") |>
    dplyr::mutate(source = stringr::str_remove_all(file, "\\.txt"),
           name = stringr::str_remove_all(name, "\\<v "), 
           comment = stringr::str_trim(comment), 
           question = stringr::str_remove_all(question, "::>")) |>
    knitr::kable()
```

+ we could change the inputs, and look at the outputs
    + so twiddle our input file, and manually check the output
+ maybe we could also change the background conditions
    + change the R environment, or package versions, or whatever
+ but that gets tedious and erratic very quickly

## `testthat`

+ Unit testing = automated, standardised, testing
+ the best place to start is with `testthat`:

```{r}
library(testthat)
```

+ [helpful man pages](https://testthat.r-lib.org/)
+ [nice vignette](https://combine-australia.github.io/r-pkg-dev/testing.html) 
+ [more ambitious guide to ad hoc testing with `testthat`](https://batteriesnotincluded.rbind.io/post/2017/08/ad-hoc-testing/)

### First steps with `testthat`

+ built for R package developers
+ but readily usable for non-package people


```{r}
test_that("multiplication works", {
  expect_equal(2 * 2, 4)
})
```

### Functions and `testthat`

+ `testthat` works best when you're testing functions
+ functions in R are easy:


```{r}
#| eval: false

function_name <- function(arg1 = default1, arg2 = default2){
     arg1 * arg2 # using our argument names
}
```



or include the body inline for simple functions:



```{r}
#| eval: false

function_name <- function(arg1 = default1, arg2 = default2) arg1 * arg2
```

### Transform your code into functions

```{r}
multo <- function(n1, n2){
  n1 * n2
}
```

### Test your function

+ then test. We think that `multo(2,2)` should equal 4, so we use:
    + `test_that()` to set up our test environment
    + `expect_equal()` inside the test environment to check for equality

```{r}
# then run as a test

test_that("multo works with 2 and 2", {
    expect_equal(multo(2, 2), 4)
})

```

### Raise your expectations

+ we can add more expectations

```{r}
test_that("multo works in general", {
    expect_equal(multo(2, 2), 4)
    expect_identical(multo(2,0.01), 0.02)
    expect_type(multo(0,2), "double")
    expect_length(multo(9,2), 1)
    expect_gte(multo(4,4), 15)
})

```

### Equal and identical

+ beware the [floating point error](https://cran.r-project.org/doc/FAQ/R-FAQ.html#Why-doesn_0027t-R-think-these-numbers-are-equal_003f)

```{r}
3 - 2.9
3 - 2.9 == 0.1
```

+ happily, there's a sufficiently sloppy way of checking equality:

```{r}
test_that("pedants corner", {
  expect_equal(multo(2, 0.01), 0.020000001)
  expect_identical(multo(2, 0.01), 0.02)
})
```

## Testing several values

+ if you want to work with vectors, there are a number of tools for checking their contents:


```{r}
x <- rownames(as.matrix(eurodist, labels=TRUE)) # odd built in dataset

test_that("check my vec", {
    expect_equal(x[1:2], c("Athens", "Barcelona"))
})    
    
```

+ you can get **much** more fancy with a bit of set theory (not really [set theory](https://plato.stanford.edu/entries/set-theory/basic-set-theory.html)):

```{r}
y <- x

test_that("check my vec sets", {
    expect_success(expect_setequal(x, y)) # all x in y
    expect_failure(expect_mapequal(x, y)) # same names, y is proper subset x) # all x in y)
    show_failure(expect_contains(x[1:19], y)) # y proper subset x)
    expect_success(expect_in(x, y)) # x proper subset y
})    
    
```


```{r}
y <- sample(x, length(x)-2)

test_that("check my vec sets", {
    expect_failure(expect_setequal(x, y)) # all x in y
    expect_failure(expect_mapequal(x, y)) # same names, y is proper subset x) # all x in y)
    expect_success(expect_contains(x, y)) # y proper subset x)
    expect_failure(expect_in(x, y)) # x is a proper subset y
})    
    
```

### Testing tibbles

+ because most of the tests are powered by [`waldo`](https://www.tidyverse.org/blog/2020/10/waldo/), you shouldn't have to do anything fancy to test on tibbles:

```{r}
library(palmerpenguins)

my_pengs <- penguins

test_that("penguin experiments", {
    expect_equal(my_pengs, penguins)
})
```

### Types and classes etc

+ one *massive* corollary to that: if you don't do a lot of base-R, expect a fiercely stringent test of your [understanding of types and classes](https://adv-r.hadley.nz/base-types.html).

```{r}
typeof(penguins)
class(penguins) 

is.object(names(penguins)) # vectors are base types
attr(names(penguins), "class") # base types have no class

is.object(penguins) # this is some kind of object
attr(penguins, "class") # so it definitely does have a class
```

### Tibble tests

```{r}

test_that("penguin types", {
    expect_type(penguins, "list")
    expect_s3_class(penguins, "tbl_df")
    expect_s3_class(penguins, "tbl")
    expect_s3_class(penguins, "data.frame")
    expect_type(penguins$island, "integer")
    expect_s3_class(penguins$island, "factor")
})

```

+ there's also an `expect_s4_class` for those with that high clear mercury sound ringing in their ears

### Last tip

+ you can put bare expectations in pipes if you're looking for something specific

```{r}
penguins |>
  expect_type("list") |>
  dplyr::pull(island) |>
  expect_length(344)
```
