---
title: "We ❤️ janitor"
author: 
  - id: 1
    name: Susanna Kirk
    email: Susanna.Kirk@nhs.scot
    affiliations: NHS Fife
  - id: 2
    name: Brendan Clarke
    email: brendan.clarke2@nhs.scot
    affiliations: NHS Education for Scotland
execute: 
  echo: true
  output: true
categories: [R, intermediate]
---

```{r}
library(tidyverse)
```


## Headline

+ Janitor is a toolkit for tidying data
+ both input-focused and output-focused
+ excellent balance between diligence and laziness

## Setup
```{r}
library(janitor)
library(readxl)

test_df <- as.data.frame(matrix(ncol = 6))
names(test_df) <- c("firstName", "ábc@!*", "% successful (2009)",
                    "REPEAT VALUE", "REPEAT VALUE", "")

test_df |>
  clean_names(case = "upper_lower") |>
  select(FIRSTname)



```

## `clean_names`

`clean_names` allows you to convert ugly column names into standardised, syntactic names. You can also specfiy the case, or bulk-tweak the names.

```{r}
clean_names(test_df)

clean_names(test_df, case = "screaming_snake")

clean_names(test_df, case = "big_camel")

test_df %>%
  clean_names(prefix = "a_prefix")

```

## `compare_df_cols`

```{r}
library(palmerpenguins)

compare_df_cols(penguins,penguins_raw)
```

## `warnings`

```{r}
#| warning: false

Data1 <- readxl::read_excel("data/JanitorDataExample.xlsx", .name_repair = make_clean_names)

# warnings switched off, otherwise we'd have literally hundreds of warnings here!

make_clean_names(c("1thing", "some opther thign"))

glimpse(Data1)
```


## `clean_names` again

```{r}
Data1_clean <- Data1 |>
  clean_names() |>
  mutate(across(contains("date"),
         excel_numeric_to_date)) |>
  mutate(Year = lubridate::year(date_of_event_1))

Data1_clean
```

## `get_dupes`

```{r}

Data1_clean |>
  get_dupes(contains("date"))

```

## `distinct`

```{r}
Data1_clean_distinct <- distinct(Data1_clean)

Data1_clean_distinct
```

## `tabyl` / `adorn`

```{r}
#| output-location: column

Data1_clean_distinct |>
  tabyl(gender)

Data1_clean_distinct |>
  tabyl(gender, show_na = F)

Data1_clean_distinct |>
  tabyl(gender, location, Year, show_na = F) # excel pivot


Data1_clean_distinct |>
  tabyl(gender) |>
  class()

Data1_clean_distinct |>
  tabyl(Year,gender, show_na = F) |>
  adorn_totals("row") |>
  adorn_percentages("row") |>
  adorn_pct_formatting() |>
  adorn_ns() |>
  adorn_title("combined")

Data1_clean_distinct |>
  tabyl(Year,gender,show_na = F) |>
  adorn_totals("row") |>
  adorn_percentages("row") |>
  adorn_pct_formatting() |>
  adorn_ns() |>
  # adorn_title("combined") |>
  mutate(Year = paste("The year was", Year))

Data1_clean_distinct |>
  tabyl(Year,gender,show_na = F) |>
  adorn_totals("row") |>
  adorn_percentages("row") |>
  adorn_pct_formatting() |>
  adorn_ns(position = "front") |>
  adorn_title("combined")
```

## group_by with `tabyl`

```{r}

Data1_clean_distinct |>
  group_by(Year,gender) |>
  summarise(Attendances = n()) |>
  filter(!is.na(Year)) |>
  pivot_wider(names_from = gender, values_from = Attendances) |>
  adorn_totals(c("row","col"))

```


## `chisq.test`

```{r}
Data1_clean_distinct |>
  tabyl(attendances,location,show_na = F) |>
  chisq.test()
```

## `chisq.test`

```{r}
titanic <- data.frame(Died  = c(122,167,528,673),
                  Survived  = c(203,118,178,212),
                  row.names = c("1st class", "2nd class", "3rd class", "crew"))

chisq.test(titanic)
```
