---
title: Writing functions in R
author:
    name: Brendan Clarke
    email: brendan.clarke2@nhs.scot
    affiliations: NHS Education for Scotland
date: 2024-06-18
execute: 
  echo: true
  output: "markup"
format: html
---


```{r}
#| results: asis

source(here::here("R/feed_block.R"))
feed_block("Writing functions in R")

```

- [slides](src/writing_functions_in_r_sl.html) / [pdf slides](src/writing_functions_in_r_sl.pdf) for this session

## Welcome!

-   this session is an 🌶🌶 **intermediate practical** designed for those with some R experience

## Session outline

+ why functions?
+ basic syntax
+ adding arguments
+ vectorised functions
+ the mystery of the paired brackets
+ ...

## Why functions?

+ most beginners write repetitious code
+ repetitious code is hard to maintain
+ functions give you an easy way of repeating chunks of code

## Basic syntax

+ think of this as a way of repeating yourself
+ in time-honoured fashion...

```{r}
hi_wrld <- function(){
  "hello world"
}

hi_wrld()
```

## Adding arguments

+ most of the time, you'll want to add **arguments** to your function
    + add a variable name inside the round bracket of `function`
    + use that variable name in your function body

```{r}
hi_wrld_n <- function(n){
  paste(rep("hello world", n))
}

hi_wrld_n(4)
```

## Another argument

+ you can add another argument
+ either position or name can be used in the function call

```{r}
hi_name_n <- function(name, n){
  rep(paste("hello", name) , n)
}

hi_name_n("sue", 4)
hi_name_n(n = 3, name = "tango") # evil but legal
```

## Defaults

```{r}
hi_name_n_def <- function(n, name = "janelle"){
  rep(paste("hello", name) , n)
}

hi_name_n_def(n = 4)
hi_name_n_def(n = 2, name = "bruce")
```

## Vectorised functions

+ most functions in R are [vectorised](https://bookdown.org/rdpeng/rprogdatascience/vectorized-operations.html)

```{r}
round(c(1.2, 3.2, 5.4, 2.7), 0)
```

+ that means that mostly, our functions will end up vectorised without us doing any work at all

```{r}
div_seven_n_round <- function(nums){
  round(nums / 7, 0)
}

numbers <- rnorm(4, 5, 50)

numbers

div_seven_n_round(numbers)
```

+ **but** there are a few cases where that can fail: most famously, using `if`/`else`

```{r}
is_even <- function(n){
  
  if(n %% 2){
  paste(n, "is odd")
} else {
  paste(n, "is even")
}
  
}
is_even(9)
is_even(10)
try(is_even(9:10))

```

## Three solutions

### vectorize with `Vectorize`

```{r}
is_even_v <- Vectorize(is_even)
is_even_v(9:10)
```

### apply

+ apply with `lapply` / `purrr::map` with `Vectorize`
```{r}
lapply(9:10, is_even)
purrr::map(9:10, is_even)
```

### refactor

+ refactor to avoid scalar functions

```{r}
is_even_rf <- function(n){
  ifelse(n %% 2, paste(n, "is odd"), paste(n, "is even"))
}
is_even_rf(9:10)
```

## `{{}}`

### What's the problem?
```{r}
mtcars |>
  dplyr::summarise(average = round(mean(hp)))
```
so

```{r}
carmo <- function(column){
  mtcars |>
    dplyr::summarise(average = round(mean(column)))
  }
```

+ but...
```{r}
try(carmo(hp))
```
> object 'hp' not found

+ we get used to R (and particularly tidyverse) helping us with some sugar when selecting column by their names
  + `mtcars$hp` / `mtcars |> select(hp)`
  + effectively, we're just able to specify `hp` like an object, and R figures out the scope etc for us
+ that misfires inside functions. R isn't sure where to look for an object called `hp`

### Enter `{{}}`

```{r}
carmo_woo <- function(column){
  mtcars |>
    dplyr::summarise(average = round(mean({{column}})))
}

carmo_woo(hp)
```

+ for 95% of purposes, take `{{}}` as a purely empirical fix
+ but, if you're very enthusiastic:
    + `{{}}` [defuses and injects](https://rlang.r-lib.org/reference/topic-metaprogramming.html) the column name
    + equivalent to `!!enquo(var)`
    
## ...

+ pass arbitrary arguments into/through a function with ...

```{r}
dotty <- function(n, ...){
  rep(paste(..., collapse = ""), n)
}

dotty(4, letters[1:5])

```


## Resources
+ **best = home made!** Refactor something simple in your code today.
+ hard to beat the treatment of functions in [R4DS](https://r4ds.hadley.nz/functions)
+ the [Rlang page on data masking](https://rlang.r-lib.org/reference/topic-data-mask.html) is surprisingly sane for such a complicated area
